---
title: "Optimization"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - r
execute:
  freeze: false
editor: 
  mode: source
---

```{r}
factors_r <- c("SP500", "DTWEXAFEGS") # "SP500" does not contain dividends; note: "DTWEXM" discontinued as of Jan 2020
factors_d <- c("DGS10", "BAMLH0A0HYM2")
```

```{r}
#| echo: false
#| output: false # Registered S3 method overwritten by 'quantmod': method as.zoo.data.frame from zoo
source("../plot/theme_jjf.R")
source("../helper-levels.R")
```

```{r}
#| echo: false
palette <- c("black", palette_jjf(length(factors)))
names(palette) <- c("Overall", factors)
```

# Random weights

Need to generate uniformly distributed weights $\mathbf{w}=(w_{1},w_{2},\ldots,w_{N})$ such that $\sum_{j=1}^{N}w_{i}=1$ and $w_{i}\geq0$:

-   **Approach 1**: tempting to use $w_{i}=\frac{u_{i}}{\sum_{j=1}^{N}u_{i}}$ where $u_{i}\sim U(0,1)$ but the distribution of $\mathbf{w}$ is not uniform

-   **Approach 2**: instead, generate $\text{Exp}(1)$ and then normalize

Can also scale random weights by $M$, e.g. if sum of weights must be 10% then multiply weights by 10%.

```{r}
rand_weights1 <- function(n_sim, n_assets) {
  
  rand_exp <- matrix(runif(n_sim * n_assets), nrow = n_sim, ncol = n_assets)
  result <- rand_exp / rowSums(rand_exp)
    
  return(result)
    
}
```

```{r}
n_assets <- 3
n_sim <- 10000
```

```{r}
approach1 <- rand_weights1(n_sim, n_assets)
```

```{r}
#| echo: false
#| fig.width: 4
#| fig.height: 4
plot_pairs(data.table::as.data.table(approach1), title = "Weight (%)")
```

**Approach 2(a)**: uniform sample from the simplex (<http://mathoverflow.net/a/76258>) and then normalize

-   If $u\sim U(0,1)$ then $-\ln(u)$ is an $\text{Exp}(1)$ distribution

This is also known as generating a random vector from the symmetric Dirichlet distribution.

```{r}
rand_weights2a <- function(n_sim, n_assets, lmbda) {
    
  # inverse transform sampling: https://en.wikipedia.org/wiki/Inverse_transform_sampling
  rand_exp <- matrix(-log(1 - runif(n_sim * n_assets)) / lmbda, nrow = n_sim, ncol = n_assets)
  result <- rand_exp / rowSums(rand_exp)
  
  return(result)
  
}
```

```{r}
lmbda <- 1
```

```{r}
approach2a <- rand_weights2a(n_sim, n_assets, lmbda)
```

```{r}
#| echo: false
#| fig.width: 4
#| fig.height: 4
plot_pairs(data.table::as.data.table(approach2a), title = "Weight (%)")
```

**Approach 2(b)**: directly generate $\text{Exp}(1)$ and then normalize

```{r}
rand_weights2b <- function(n_sim, n_assets) {
  
  rand_exp <- matrix(rexp(n_sim * n_assets), nrow = n_sim, ncol = n_assets)
  result <- rand_exp / rowSums(rand_exp)
  
  return(result)
  
}
```

```{r}
approach2b <- rand_weights2b(n_sim, n_assets)
```

```{r}
#| echo: false
#| fig.width: 4
#| fig.height: 4
plot_pairs(data.table::as.data.table(approach2b), title = "Weight (%)")
```

## Random turnover

How to generate random weights between lower bound $a$ and upper bound $b$ that sum to zero?

-   **Approach 1**: tempting to multiply random weights by $M$ and then subtract by $\frac{M}{N}$ but the distribution is not between $a$ and $b$

-   **Approach 2**: instead, use an iterative approach for random turnover:

    1.  Generate $N-1$ uniformly distributed weights between $a$ and $b$
    2.  For $u_{N}$ compute sum of values and subtract from $M$
    3.  If $u_{N}$ is between $a$ and $b$, then keep; otherwise, discard

Then add random turnover to previous period's random weights.

```{r}
rand_turnover1 <- function(n_sim, n_assets, lower, upper, target) {
    
  rng <- upper - lower
  
  result <- rand_weights2b(n_sim, n_assets) * rng
  result <- result - rng / n_assets
  
  return(result)
    
}
```

```{r}
lower <- -0.05
upper <- 0.05
target <- 0
```

```{r}
approach1 <- rand_turnover1(n_sim, n_assets, lower, upper, target)
```

```{r}
#| echo: false
#| fig.width: 4
#| fig.height: 4
plot_pairs(data.table::as.data.table(approach1), title = "Weight (%)")
```

```{r}
rand_iterative <- function(n_assets, lower, upper, target) {
  
  result <- runif(n_assets - 1, min = lower, max = upper)
  temp <- target - sum(result)
  
  while (!((temp <= upper) && (temp >= lower))) {
      
    result <- runif(n_assets - 1, min = lower, max = upper)
    temp <- target - sum(result)
      
  }
  
  result <- append(result, temp)
  
  return(result)
  
}
```

```{r}
rand_turnover2 <- function(n_sim, n_assets, lower, upper, target) {
  
  result_ls <- list()
  
  for (i in 1:n_sim) {
      
    result_sim <- rand_iterative(n_assets, lower, upper, target)
    result_ls <- append(result_ls, list(result_sim))
      
  }
  
  result <- do.call(rbind, result_ls)
  
  return(result)
  
}
```

```{r}
approach2 <- rand_turnover2(n_sim, n_assets, lower, upper, target)
```

```{r}
#| echo: false
#| fig.width: 4
#| fig.height: 4
plot_pairs(data.table::as.data.table(approach2), title = "Weight (%)")
```

# Mean-variance

```{r}
geometric_mean <- function(x, scale) {
  
  result <- prod(1 + x) ^ (scale / length(x)) - 1
  
  return(result)
  
}
```

-   <https://www.adrian.idv.hk/2021-06-22-kkt/>
-   <https://or.stackexchange.com/a/3738>
-   <https://bookdown.org/compfinezbook/introFinRbook/Portfolio-Theory-with-Matrix-Algebra.html#algorithm-for-computing-efficient-frontier>
-   <https://palomar.home.ece.ust.hk/MAFS6010R_lectures/slides_robust_portfolio.html>

```{r}
tickers <- "BAICX" # fund inception date is "2011-11-28" 
```

```{r}
#| echo: false
source("../helper-prices-yfh.R")

overlap_xts <- roll::roll_mean(returns_xts, scale[["overlap"]], min_obs = 1)

overlap_xts <- na.omit(overlap_xts)
overlap_x_xts <- overlap_xts[ , factors]
overlap_y_xts <- overlap_xts[ , tickers]
```

```{r}
returns_x_xts <- na.omit(returns_xts)[ , factors] # extended history
mu <- apply(returns_x_xts, 2, geometric_mean, scale = scale[["periods"]])
sigma <- cov(overlap_x_xts) * scale[["periods"]] * scale[["overlap"]]
```

## Maximize mean

$$
\begin{aligned}
\begin{array}{rrcl}
\displaystyle\min&-\mathbf{w}^{T}\mu\\
\textrm{s.t.}&\mathbf{w}^{T}e&=&1\\
&\mathbf{w}^T\Sigma\mathbf{w}&\leq&\sigma^{2}\\
\end{array}
\end{aligned}
$$

To incorporate these conditions into one equation, introduce new variables $\lambda_{i}$ that are the Lagrange multipliers and define a new function $\mathcal{L}$ as follows:

$$
\begin{aligned}
\mathcal{L}(\mathbf{w},\lambda)&=-\mathbf{w}^{T}\mu-\lambda_{1}(\mathbf{w}^{T}e-1)
\end{aligned}
$$

Then, to minimize this function, take derivatives with respect to $w$ and Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial w}&=-\mu-\lambda_{1}e=0\\
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial \lambda_{1}}&=\mathbf{w}e^T-1=0
\end{aligned}
$$

Simplify the equations above in matrix form and solve for the Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\begin{bmatrix}
-\mu & e \\
e^{T} & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
0 \\
1
\end{bmatrix}
\\
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
-\mu & e \\
e^{T} & 0
\end{bmatrix}^{-1}
\begin{bmatrix}
0 \\
1
\end{bmatrix}
\end{aligned}
$$

```{r}
max_mean_optim <- function(mu, sigma, target) {
  
  params <- CVXR::Variable(length(mu))
  
  obj <- CVXR::Maximize(t(params) %*% mu)
  
  cons <- list(sum(params) == 1, params >= 0,
               CVXR::quad_form(params, sigma) <= target ^ 2)
  
  prob <- CVXR::Problem(obj, cons)
  
  result <- CVXR::solve(prob)$getValue(params)
  
  return(result)

}
```

```{r}
target <- 0.06
```

```{r}
params1 <- t(max_mean_optim(mu, sigma, target))
params1
```

```{r}
params1 %*% mu
```

```{r}
sqrt(params1 %*% sigma %*% t(params1))
```

```{r}
# # install.packages("devtools")
# devtools::install_github("jasonjfoster/rolloptim") # roll (>= 1.1.7)
# library(rolloptim)
# 
# mu <- roll_mean(returns_x_xts, 5)
# sigma <- roll_cov(returns_x_xts, width = 5)
# 
# xx <- roll_crossprod(returns_x_xts, returns_x_xts, 5)
# xy <- roll_crossprod(returns_x_xts, returns_y_xts, 5)
# 
# roll_max_mean(mu)
```

## Minimize variance

$$
\begin{aligned}
\begin{array}{rrcl}
\displaystyle\min&\frac{1}{2}\mathbf{w}^T\Sigma\mathbf{w}\\
\textrm{s.t.}&\mathbf{w}^{T}e&=&1\\
&\mu^{T}\mathbf{w}&\geq&M\\
\end{array}
\end{aligned}
$$

To incorporate these conditions into one equation, introduce new variables $\lambda_{i}$ that are the Lagrange multipliers and define a new function $\mathcal{L}$ as follows:

$$
\begin{aligned}
\mathcal{L}(\mathbf{w},\lambda)&=\frac{1}{2}\mathbf{w}^{T}\Sigma\mathbf{w}-\lambda_{1}(\mathbf{w}^{T}e-1)
\end{aligned}
$$

Then, to minimize this function, take derivatives with respect to $w$ and Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial w}&=\mathbf{w}\Sigma-\lambda_{1}e=0\\
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial \lambda_{1}}&=\mathbf{w}e^T-1=0
\end{aligned}
$$

Simplify the equations above in matrix form and solve for the Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\begin{bmatrix}
\Sigma & e \\
e^{T} & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
0 \\
1
\end{bmatrix}
\\
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
\Sigma & e \\
e^{T} & 0
\end{bmatrix}^{-1}
\begin{bmatrix}
0 \\
1
\end{bmatrix}
\end{aligned}
$$

```{r}
min_var_optim <- function(mu, sigma, target) {
  
  params <- CVXR::Variable(length(mu))
  
  obj <- CVXR::Minimize(CVXR::quad_form(params, sigma))
  
  cons <- list(sum(params) == 1, params >= 0,
               sum(mu * params) >= target)
  
  prob <- CVXR::Problem(obj, cons)
  
  result <- CVXR::solve(prob)$getValue(params)
  
  return(result)

}
```

```{r}
target <- 0.03
```

```{r}
params2 <- t(min_var_optim(mu, sigma, target))
params2
```

```{r}
params2 %*% mu
```

```{r}
sqrt(params2 %*% sigma %*% t(params2))
```

```{r}
# roll_min_var(sigma)
```

## Maximize utility

$$
\begin{aligned}
\begin{array}{rrcl}
\displaystyle\min&\frac{1}{2}\delta(\mathbf{w}^{T}\Sigma\mathbf{w})-\mu^{T}\mathbf{w}\\
\textrm{s.t.}&\mathbf{w}^{T}e&=&1\\
\end{array}
\end{aligned}
$$

To incorporate these conditions into one equation, introduce new variables $\lambda_{i}$ that are the Lagrange multipliers and define a new function $\mathcal{L}$ as follows:

$$
\begin{aligned}
\mathcal{L}(\mathbf{w},\lambda)&=\frac{1}{2}\mathbf{w}^{T}\Sigma\mathbf{w}-\mu^{T}\mathbf{w}-\lambda_{1}(\mathbf{w}^{T}e-1)
\end{aligned}
$$

Then, to minimize this function, take derivatives with respect to $w$ and Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial w}&=\mathbf{w}\Sigma-\mu^{T}-\lambda_{1}e=0\\
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial \lambda_{1}}&=\mathbf{w}e^T-1=0
\end{aligned}
$$

Simplify the equations above in matrix form and solve for the Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\begin{bmatrix}
\Sigma & e \\
e^{T} & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
\mu^{T} \\
1
\end{bmatrix}
\\
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
\Sigma & e \\
e^{T} & 0
\end{bmatrix}^{-1}
\begin{bmatrix}
\mu^{T} \\
1
\end{bmatrix}
\end{aligned}
$$

```{r}
max_utility_optim <- function(mu, sigma, target) {
  
  params <- CVXR::Variable(length(mu))
  
  obj <- CVXR::Minimize(0.5 * target * CVXR::quad_form(params, sigma) - t(mu) %*% params)
  
  cons <- list(sum(params) == 1, params >= 0)
  
  prob <- CVXR::Problem(obj, cons)
      
  result <- CVXR::solve(prob)$getValue(params)
  
  return(result)

}
```

```{r}
ir <- 0.5
target <- ir / 0.06 # ir / std (see Black-Litterman)
```

```{r}
params3 <- t(max_utility_optim(mu, sigma, target))
params3
```

```{r}
params3 %*% mu 
```

```{r}
sqrt(params3 %*% sigma %*% t(params3))
```

```{r}
# roll_max_utility(mu, sigma)
```

## Minimize residual sum of squares

$$
\begin{aligned}
\begin{array}{rrcl}
\displaystyle\min&\frac{1}{2}\delta(\mathbf{w}^{T}X^{T}X\mathbf{w})-X^{T}y\mathbf{w}\\
\textrm{s.t.}&\mathbf{w}^{T}e&=&1\\
\end{array}
\end{aligned}
$$

To incorporate these conditions into one equation, introduce new variables $\lambda_{i}$ that are the Lagrange multipliers and define a new function $\mathcal{L}$ as follows:

$$
\begin{aligned}
\mathcal{L}(\mathbf{w},\lambda)&=\frac{1}{2}\mathbf{w}^{T}X^{T}X\mathbf{w}-X^{T}y\mathbf{w}-\lambda_{1}(\mathbf{w}^{T}e-1)
\end{aligned}
$$

Then, to minimize this function, take derivatives with respect to $w$ and Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial w}&=\mathbf{w}X^{T}X-X^{T}y-\lambda_{1}e=0\\
\frac{\partial\mathcal{L}(\mathbf{w},\lambda)}{\partial \lambda_{1}}&=\mathbf{w}e^T-1=0
\end{aligned}
$$

Simplify the equations above in matrix form and solve for the Lagrange multipliers $\lambda_{i}$:

$$
\begin{aligned}
\begin{bmatrix}
X^{T}X & e \\
e^{T} & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
X^{T}y \\
1
\end{bmatrix}
\\
\begin{bmatrix}
\mathbf{w} \\
-\lambda_{1}
\end{bmatrix}
&=
\begin{bmatrix}
X^{T}X & e \\
e^{T} & 0
\end{bmatrix}^{-1}
\begin{bmatrix}
X^{T}y \\
1
\end{bmatrix}
\end{aligned}
$$

-   <https://scaron.info/blog/conversion-from-least-squares-to-quadratic-programming.html>

```{r}
min_rss_optim1 <- function(mu, sigma) {
    
  params <- CVXR::Variable(length(mu))
  
  obj <- CVXR::Minimize(0.5 * CVXR::quad_form(params, sigma) - t(mu) %*% params)
  
  cons <- list(sum(params) == 1, params >= 0)
  
  prob <- CVXR::Problem(obj, cons)
      
  result <- CVXR::solve(prob)$getValue(params)
  
  return(result)

}
```

```{r}
params4 <- t(min_rss_optim1(crossprod(overlap_x_xts, overlap_y_xts), crossprod(overlap_x_xts)))
params4
```

```{r}
params4 %*% mu 
```

```{r}
sqrt(params4 %*% sigma %*% t(params4))
```

```{r}
# roll_min_rss(xx, xy)
```

```{r}
min_rss_optim2 <- function(x, y) {
  
  params <- CVXR::Variable(ncol(x))
  
  obj <- CVXR::Minimize(CVXR::sum_squares(y - x %*% params))
  
  cons <- list(sum(params) == 1, params >= 0)
  
  prob <- CVXR::Problem(obj, cons)
      
  result <- CVXR::solve(prob)$getValue(params)
  
  return(result)

}
```

```{r}
params5 <- t(min_rss_optim2(zoo::coredata(overlap_x_xts), zoo::coredata(overlap_y_xts)))
params5
```

```{r}
params5 %*% mu 
```

```{r}
sqrt(params5 %*% sigma %*% t(params5))
```

```{r}
round(data.frame("max_pnl" = t(params1) * 100,
                 "min_risk" = t(params2) * 100,
                 "max_utility" = t(params3) * 100,
                 "min_rss1" = t(params4) * 100,
                 "min_rss2" = t(params5) * 100), 2)
```
