---
title: "Markets"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - python
execute:
  freeze: false
editor: 
  mode: source
---

```{python}
factors_r = ["SP500", "DTWEXAFEGS"] # "SP500" does not contain dividends; note: "DTWEXM" discontinued as of Jan 2020
factors_d = ["DGS10", "BAMLH0A0HYM2"]
```

```{r}
#| echo: false
source("../plot/theme_jjf.R")
```

```{python}
#| echo: false
exec(open("../helper-levels.py").read())
```

```{r}
#| echo: false
palette <- c("black", palette_jjf(length(reticulate::py$factors)))
names(palette) <- c("Overall", reticulate::py$factors)
```

# Price momentum

One month reversal and 2-12 month momentum are two ends of the spectrum. The general trend indicates that positive acceleration leads to reversals or negative acceleration leads to rebounds. An unsustainable acceleration leading to reversal can reconcile the one-month reversal and 2-12 month momentum. The key is that it implies that acceleration is not sustainable.

```{python}
# "Momentum, Acceleration, and Reversal"
def pnl(x):
  return np.nanprod(1 + x) - 1
```

```{python}
order = 20
```

```{python}
momentum_df = returns_df.shift(order).rolling(width - order, min_periods = 1).apply(pnl, raw = False).dropna()
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
momentum_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$momentum_df, keep.rownames = "index"), id.vars = "index")
momentum_mlt[ , index := as.Date(index)]
momentum_plt <- plot_ts_decomp(momentum_mlt, decomp = "Total", title = "Momentum 1Y (%)") +
  facet_wrap(~ variable, scales = "free")
print(momentum_plt)
```

# Time-series score

Suppose we are looking at $n$ independent and identically distributed random variables, $X_{1},X_{2},\ldots,X_{n}$. Since they are iid, each random variable $X_{i}$ has to have the same mean, which we will call $\mu$, and variance, which we will call $\sigma^{2}$:

$$
\begin{aligned}
\mathrm{E}\left(X_{i}\right)&=\mu\\
\mathrm{Var}\left(X_{i}\right)&=\sigma^{2}
\end{aligned}
$$

Let's suppose we want to look at the average value of our $n$ random variables:

$$
\begin{aligned}
\bar{X}=\frac{X_{1}+X_{2}+\cdots+X_{n}}{n}=\left(\frac{1}{n}\right)\left(X_{1}+X_{2}+\cdots+X_{n}\right)
\end{aligned}
$$

We want to find the expected value and variance of the average, $\mathrm{E}\left(\bar{X}\right)$ and $\mathrm{Var}\left(\bar{X}\right)$.

## Expected value

$$
\begin{aligned}
\mathrm{E}\left(\bar{X}\right)&=\mathrm{E}\left[\left(\frac{1}{n}\right)\left(X_{1}+X_{2}+\cdots+X_{n}\right)\right]\\
&=\left(\frac{1}{n}\right)\mathrm{E}\left(X_{1}+X_{2}+\cdots+X_{n}\right)\\
&=\left(\frac{1}{n}\right)\left(n\mu\right)\\
&=\mu
\end{aligned}
$$

## Variance

$$
\begin{aligned}
\mathrm{Var}\left(\bar{X}\right)&=\mathrm{Var}\left[\left(\frac{1}{n}\right)\left(X_{1}+X_{2}+\cdots+X_{n}\right)\right]\\
&=\left(\frac{1}{n}\right)^{2}\mathrm{Var}\left(X_{1}+X_{2}+\cdots+X_{n}\right)\\
&=\left(\frac{1}{n}\right)^{2}\left(n\sigma^{2}\right)\\
&=\frac{\sigma^{2}}{n}
\end{aligned}
$$

-   <http://scipp.ucsc.edu/~haber/ph116C/iid.pdf>

```{python}
def sd(x):
    
  n_rows = sum(~np.isnan(x))
      
  if n_rows > 1:
    result = np.sqrt(np.nansum(x ** 2) / (n_rows - 1))
  else:
    result = np.nan
      
  return result
```

```{python}
# volatility scale only
score_df = (momentum_df / momentum_df.rolling(width, min_periods = 1).apply(sd, raw = False)).dropna()
```

```{python}
# overall_df = score_df.mean(axis = 1)
# overall_df = overall_df / overall_df.rolling(width, min_periods = 1).apply(risk, raw = False)
```

```{python}
# score_df.insert(loc = 0, column = "Overall", value = overall_df)
# score_df = score_df.dropna()
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
score_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$score_df, keep.rownames = "index"), id.vars = "index")
score_mlt[ , index := as.Date(index)]
score_plt <- plot_ts_decomp(score_mlt, decomp = "Overall", title = "Score 1Y", multiple = 1) +
  facet_wrap(~ variable)
print(score_plt)
```

# Outlier detection

```{python}
import statsmodels.api as sm
```

## Interquartile range

Outliers are defined as the regression residuals that fall below $Q_{1}−1.5\times IQR$ or above $Q_{3}+1.5\times IQR$:

-   <https://stats.stackexchange.com/a/1153>
-   <https://stats.stackexchange.com/a/108951>
-   <https://robjhyndman.com/hyndsight/tsoutliers/>

```{python}
def outliers(z):
  
  n_cols = z.shape[1]
  result_ls = []

  for j in range(n_cols):
    
    y = z.iloc[:, j]

    if (n_cols == 0):
      x = sm.add_constant(range(len(y)))
    else:
      x = sm.add_constant(z.drop(z.columns[j], axis = 1))

    coef = sm.WLS(y, x).fit().params
    predict = coef.iloc[0] + np.dot(x.iloc[:, 1:], coef[1:])
    resid = y - predict

    lower = resid.quantile(0.25)
    upper = resid.quantile(0.75)
    iqr = upper - lower

    total = y[(resid < lower - 1.5 * iqr) | (resid > upper + 1.5 * iqr)]
    
    total = pd.DataFrame({"date": total.index, "symbol": total.name, "values": total})
    result_ls.append(total)

  result = pd.concat(result_ls, ignore_index = True)
  result = result.pivot_table(index = "date", columns = "symbol", values = "values")

  return result
```

```{python}
outliers_df = outliers(score_df)
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
outliers_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$outliers_df, keep.rownames = "index"), id.vars = "index")
outliers_mlt[ , index := as.Date(index)]
outliers_plt <- score_plt +
  geom_point(data = outliers_mlt, aes(x = index, y = value, fill = variable))
print(outliers_plt)
```

## Contour ellipsoid

# Granger causality

```{python}
from scipy.stats import chi2
```

$$
\begin{aligned}
\left(R\hat{\beta}-r\right)^\mathrm{T}\left(R\hat{V}R^\mathrm{T}\right)^{-1}\left(R\hat{\beta}-r\right)\xrightarrow\quad\chi_{Q}^{2}
\end{aligned}
$$

-   <https://github.com/cran/lmtest/blob/master/R/waldtest.R>
-   <https://en.wikipedia.org/wiki/Wald_test#Test(s)_on_multiple_parameters>
-   <https://math.stackexchange.com/a/1591946>

```{python}
def granger_test(x, y, order):
    
  # compute lagged observations
  lag_x = x.shift(order)
  lag_y = y.shift(order)
  
  # collect series
  df = pd.DataFrame({"x": x, "y": y, "lag_x": lag_x, "lag_y": lag_y})
  x = sm.add_constant(df[["lag_y", "lag_x"]])
  
  # fit full model
  fit = sm.WLS(df["y"], x, missing = "drop").fit()
  
  R = np.array([0, 0, 1])
  coef = fit.params
  r = 0 # technically a matrix (see Stack Exchange)
  
  matmul = np.dot(R, coef) - r
  matmul_mid = np.linalg.inv(np.atleast_2d(np.dot(R, np.dot(fit.cov_params(), R.T))))
  wald = np.dot(matmul.T, np.dot(matmul_mid, matmul))
  
  result = 1 - chi2.cdf(wald, 1)
  
  return np.ravel(result)
```

```{python}
def roll_lead_lag(x, y, width, order, p_value):
  
  n_rows = len(x)
  x_name = x.name
  y_name = y.name
  x_y_ls = []
  y_x_ls = []

  for i in range(width - 1, n_rows):
    
    idx = range(max(i - width + 1, 0), i + 1)

    x_y = granger_test(x.iloc[idx], y.iloc[idx], order)
    y_x = granger_test(y.iloc[idx], x.iloc[idx], order)

    x_y_status = (x_y < p_value) and (y_x > p_value)
    y_x_status = (x_y > p_value) and (y_x < p_value)
    
    x_y_ls.append(x_y_status)
    y_x_ls.append(y_x_status)
  
  result = pd.DataFrame({x_name: x_y_ls, y_name: y_x_ls}, index = x.index[(width - 1):])

  return result
```

```{python}
p_value = 0.05
```

```{python}
score_x_df = score_df.loc[:, "SP500"]
score_y_df = score_df.loc[:, "DGS10"]
```

```{python}
lead_lag_df = roll_lead_lag(score_x_df, score_y_df, width, order, p_value)
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
lead_lag_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$lead_lag_df, keep.rownames = "index"), id.vars = "index")
lead_lag_mlt[ , index := as.Date(index)]
lead_lag_mlt[ , value := as.logical(value)]
lead_lag_plt <- plot_ts_decomp(lead_lag_mlt, decomp = "Total", title = "Lead lag 1M", palette = palette) +
  theme(axis.text.y = element_blank())
print(lead_lag_plt)
```
