---
title: "Markets"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - python
execute:
  freeze: false
editor: 
  mode: source
---

```{python}
factors_r = ["SP500", "DTWEXAFEGS"] # "SP500" does not contain dividends; note: "DTWEXM" discontinued as of Jan 2020
factors_d = ["DGS10", "BAMLH0A0HYM2"]
```

```{r}
#| echo: false
source("../plot/theme_jjf.R")
```

```{python}
#| echo: false
exec(open("../helper-levels.py").read())
```

```{r}
#| echo: false
palette <- c("black", palette_jjf(length(reticulate::py$factors)))
names(palette) <- c("Overall", reticulate::py$factors)
```

# Price momentum

One month reversal and 2-12 month momentum are two ends of the spectrum. The general trend indicates that positive acceleration leads to reversals and negative acceleration leads to rebounds, i.e. acceleration is not sustainable. An unsustainable acceleration leading to reversal can reconcile the one-month reversal and 2-12 month momentum.

```{python}
order = 20
```

```{python}
# "Momentum, Acceleration, and Reversal"
momentum_df = (np.exp(returns_df.shift(order).rolling(width - order, min_periods = 1).sum()) - 1).dropna()
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
momentum_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$momentum_df, keep.rownames = "index"), id.vars = "index")
momentum_mlt[ , index := as.Date(index)]
momentum_plt <- plot_ts_decomp(momentum_mlt, decomp = "Total", title = "Momentum 1Y (%)") +
  ggplot2::facet_wrap(~ variable, scales = "free")
print(momentum_plt)
```

# Time-series score

Suppose we are looking at $n$ independent and identically distributed random variables, $X_{1},X_{2},\ldots,X_{n}$. Since they are iid, each random variable $X_{i}$ has to have the same mean, which we will call $\mu$, and variance, which we will call $\sigma^{2}$:

$$
\begin{aligned}
\mathrm{E}\left(X_{i}\right)&=\mu\\
\mathrm{Var}\left(X_{i}\right)&=\sigma^{2}
\end{aligned}
$$

Let's suppose we want to look at the average value of our $n$ random variables:

$$
\begin{aligned}
\bar{X}=\frac{X_{1}+X_{2}+\cdots+X_{n}}{n}=\left(\frac{1}{n}\right)\left(X_{1}+X_{2}+\cdots+X_{n}\right)
\end{aligned}
$$

We want to find the expected value and variance of the average, $\mathrm{E}\left(\bar{X}\right)$ and $\mathrm{Var}\left(\bar{X}\right)$.

## Expected value

$$
\begin{aligned}
\mathrm{E}\left(\bar{X}\right)&=\mathrm{E}\left[\left(\frac{1}{n}\right)\left(X_{1}+X_{2}+\cdots+X_{n}\right)\right]\\
&=\left(\frac{1}{n}\right)\mathrm{E}\left(X_{1}+X_{2}+\cdots+X_{n}\right)\\
&=\left(\frac{1}{n}\right)\left(n\mu\right)\\
&=\mu
\end{aligned}
$$

## Variance

$$
\begin{aligned}
\mathrm{Var}\left(\bar{X}\right)&=\mathrm{Var}\left[\left(\frac{1}{n}\right)\left(X_{1}+X_{2}+\cdots+X_{n}\right)\right]\\
&=\left(\frac{1}{n}\right)^{2}\mathrm{Var}\left(X_{1}+X_{2}+\cdots+X_{n}\right)\\
&=\left(\frac{1}{n}\right)^{2}\left(n\sigma^{2}\right)\\
&=\frac{\sigma^{2}}{n}
\end{aligned}
$$

-   <http://scipp.ucsc.edu/~haber/ph116C/iid.pdf>

```{python}
def sd(x):
    
  n_rows = sum(~np.isnan(x))
      
  if n_rows > 1:
    result = np.sqrt(np.nansum(x ** 2) / (n_rows - 1))
  else:
    result = np.nan
      
  return result
```

```{python}
# volatility scale only
score_df = (momentum_df / momentum_df.rolling(width, min_periods = 1).apply(sd, raw = False)).dropna()
```

```{python}
# overall_df = score_df.mean(axis = 1)
# overall_df = overall_df / overall_df.rolling(width, min_periods = 1).apply(sd, raw = False)
```

```{python}
# score_df.insert(loc = 0, column = "Overall", value = overall_df)
# score_df = score_df.dropna()
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
score_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$score_df, keep.rownames = "index"), id.vars = "index")
score_mlt[ , index := as.Date(index)]
score_plt <- plot_ts_decomp(score_mlt, decomp = "Overall", title = "Score 1Y", multiple = 1) +
  ggplot2::facet_wrap(~ variable)
print(score_plt)
```

# Outlier detection

```{python}
import statsmodels.api as sm
```

## Interquartile range

Outliers are defined as the regression residuals that fall below $Q_{1}-1.5\times IQR$ or above $Q_{3}+1.5\times IQR$:

-   <https://stats.stackexchange.com/a/1153>
-   <https://stats.stackexchange.com/a/108951>
-   <https://robjhyndman.com/hyndsight/tsoutliers/>

```{python}
def outliers(z):
  
  n_cols = z.shape[1]
  result_ls = []

  for j in range(n_cols):
    
    y = z.iloc[:, j]

    trend = pd.DataFrame({"trend": range(len(y))}, index = y.index)

    if (n_cols == 1):
      x = sm.add_constant(trend)
    else:
      x = sm.add_constant(pd.concat([trend, z.drop(z.columns[j], axis = 1)], axis = 1))

    coef = sm.OLS(y, x).fit().params
    predict = coef.iloc[0] + np.dot(x.iloc[:, 1:], coef.iloc[1:])
    resid = y - predict

    lower = resid.quantile(0.25)
    upper = resid.quantile(0.75)
    iqr = upper - lower

    total = y[(resid < lower - 1.5 * iqr) | (resid > upper + 1.5 * iqr)]
    
    total = pd.DataFrame({"date": total.index, "symbol": total.name, "values": total})
    result_ls.append(total)

  result = pd.concat(result_ls, ignore_index = True)
  result = result.pivot_table(index = "date", columns = "symbol", values = "values")

  return result
```

```{python}
outliers_df = outliers(score_df)
```

```{r}
#| echo: false
#| warning: false # Removed rows containing missing values or values outside the scale range (`geom_point()`)
#| fig.width: 6
#| fig.height: 3
outliers_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$outliers_df, keep.rownames = "index"), id.vars = "index")
outliers_mlt[ , index := as.Date(index)]
outliers_plt <- score_plt +
  ggplot2::geom_point(
    data = outliers_mlt,
    ggplot2::aes(x = index, y = value, fill = variable)
  )
print(outliers_plt)
```

## Contour ellipsoid

The contours of a multivariate normal (MVN) distribution are ellipsoids centered at the mean. The directions of the axes are given by the eigenvectors of the covariance matrix and squared lengths are given by the eigenvalues:

$$
\begin{aligned}
({\mathbf{x}}-{\boldsymbol{\mu}})^{\mathrm{T}}{\boldsymbol{\Sigma}}^{-1}({\mathbf{x}}-{\boldsymbol{\mu}})=c^{2}
\end{aligned}
$$

Or, in general parametric form:

$$
\begin{aligned}
X(t)&=X_{c}+a\,\cos t\,\cos \varphi -b\,\sin t\,\sin \varphi\\
Y(t)&=Y_{c}+a\,\cos t\,\sin \varphi +b\,\sin t\,\cos \varphi
\end{aligned}
$$ where $t$ varies from $0,\ldots,2\pi$. Here $(X_{c},Y_{c})$ is the center of the ellipse and $\varphi$ is the angle between the x-axis and the major axis of the ellipse.

Specifically:

$$
\begin{aligned}
&\text{Center: }\boldsymbol{\mu}=(X_{c},Y_{c})\\
&\text{Radius: }c^{2}= \chi_{\alpha}^{2}(df)\\
&\text{Length: }a=c\sqrt{\lambda_{k}}\\
&\text{Angle of rotation: }\varphi=\text{atan2}\left(V_{k}(2),V_{k}(1)\right)
\end{aligned}
$$

Note: the radius converts a normal quantile into a chi-squared quantile, i.e. $c^{2}=\chi_{\alpha}^{2}(2)$ with $\alpha=2\Phi(\sigma)-1$, so a one-sigma ellipse covers $2\Phi(1)-1\approx68\%$ of the distribution, i.e. the familiar two-sided normal coverage.

```{python}
# https://www.visiondummy.com/2014/04/draw-error-ellipse-representing-covariance-matrix/
# https://maitra.public.iastate.edu/stat501/lectures/MultivariateNormalDistribution-I.pdf
# https://en.wikipedia.org/wiki/Multivariate_normal_distribution
# https://en.wikipedia.org/wiki/Ellipse#General_parametric_form
from scipy.stats import chi2, norm

def ellipse(n_sim, x, y, sigma):

    data = np.concatenate((x, y), axis = 1)
    L, V = np.linalg.eig(np.cov(data.T, ddof = 1))
    idx = L.argsort()[::-1]
    L = L[idx]
    V = V[:, idx]

    c = np.sqrt(chi2.ppf(2 * norm.cdf(sigma) - 1, 2))
    t = np.linspace(0, 2 * np.pi, n_sim)
    phi = np.arctan2(V[1, 0], V[0, 0])
    a = c * np.sqrt(L[0]) * np.cos(t)
    b = c * np.sqrt(L[1]) * np.sin(t)
    R = np.array([[np.cos(phi), np.sin(phi)], [-np.sin(phi), np.cos(phi)]])
    r = np.matmul(np.column_stack([a, b]), R)

    result = np.add(r, np.mean(data, axis = 0)) # 2D only

    return result
```

```{python}
returns_x_df = returns_df.dropna()[factors] # extended history
returns_x_mat = np.asarray(returns_x_df)
ellipse_x_mat = ellipse(1000, returns_x_mat[:, [0]], returns_x_mat[:, [2]], 1)
```

```{r}
#| echo: false
#| fig.width: 3
#| fig.height: 3
ellipse_plt <- plot_scatter(data.table::data.table(reticulate::py$returns_x_mat[ , c(1, 3)]), x = "V1", y = "V2",
                            title = "Return 1D (%)", xlab = "SP500", ylab = "DGS10") +
  ggplot2::geom_point(data = data.table::data.table(reticulate::py$ellipse_x_mat), ggplot2::aes(x = V1 * 100, y = V2 * 100))
print(ellipse_plt)
```

# Granger causality

Granger causality tests whether lagged values of one series improve predictions of another series. Regress each series on lagged values of itself and the other series, then use a Wald test on the coefficient of the other series, i.e. the restriction $R\hat{\beta}=r$ (*note: condition on all own lags up to the momentum order but test a single cross lag at the momentum order*):

$$
\begin{aligned}
\left(R\hat{\beta}-r\right)^\mathrm{T}\left(R\hat{V}R^\mathrm{T}\right)^{-1}\left(R\hat{\beta}-r\right)\xrightarrow{d}\chi_{Q}^{2}
\end{aligned}
$$

-   <https://github.com/cran/lmtest/blob/master/R/waldtest.R>
-   <https://en.wikipedia.org/wiki/Wald_test#Test(s)_on_multiple_parameters>
-   <https://math.stackexchange.com/a/1591946>

```{python}
def granger_test(x, y, order):

  # compute lagged observations
  lag_x = x.shift(order).rename("lag_x")
  lag_y = pd.concat([y.shift(i) for i in range(1, order + 1)], axis = 1)
  lag_y.columns = ["lag_y" + str(i) for i in range(1, order + 1)]

  # collect series
  df = pd.concat([y.rename("y"), lag_y, lag_x], axis = 1)
  z = sm.add_constant(df.drop("y", axis = 1))

  # fit full model
  fit = sm.OLS(df["y"], z, missing = "drop").fit()

  R = np.append(np.zeros(order + 1), 1)
  coef = fit.params
  r = 0 # technically a matrix (see Stack Exchange)

  matmul = np.dot(R, coef) - r
  matmul_mid = np.linalg.inv(np.atleast_2d(np.dot(R, np.dot(fit.cov_params(), R.T))))
  wald = np.dot(matmul.T, np.dot(matmul_mid, matmul))

  result = 1 - chi2.cdf(wald, 1)

  return result
```

```{python}
def roll_lead_lag(x, y, width, order, sig_level):
  
  n_rows = len(x)
  x_name = x.name
  y_name = y.name
  x_y_ls = []
  y_x_ls = []

  for i in range(width - 1, n_rows):
    
    idx = range(max(i - width + 1, 0), i + 1)

    x_y = granger_test(x.iloc[idx], y.iloc[idx], order)
    y_x = granger_test(y.iloc[idx], x.iloc[idx], order)

    x_y_status = (x_y < sig_level) and (y_x > sig_level)
    y_x_status = (x_y > sig_level) and (y_x < sig_level)
    
    x_y_ls.append(x_y_status)
    y_x_ls.append(y_x_status)
  
  result = pd.DataFrame({x_name: x_y_ls, y_name: y_x_ls}, index = x.index[(width - 1):])

  return result
```

```{python}
sig_level = 0.05
```

```{python}
score_x_df = score_df.loc[:, "SP500"]
score_y_df = score_df.loc[:, "DGS10"]
```

```{python}
lead_lag_df = roll_lead_lag(score_x_df, score_y_df, width, order, sig_level)
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
lead_lag_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$lead_lag_df, keep.rownames = "index"), id.vars = "index")
lead_lag_mlt[ , index := as.Date(index)]
lead_lag_mlt[ , value := as.logical(value)]
lead_lag_plt <- plot_ts_decomp(lead_lag_mlt, decomp = "Total", title = "Lead lag 1Y", palette = palette) +
  ggplot2::theme(axis.text.y = ggplot2::element_blank())
print(lead_lag_plt)
```
