---
title: "Eigen"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - python
execute:
  freeze: false
editor: 
  mode: source
---

```{python}
factors_r = ["SP500", "DTWEXAFEGS"] # "SP500" does not contain dividends; note: "DTWEXM" discontinued as of Jan 2020
factors_d = ["DGS10", "BAMLH0A0HYM2"]
```

```{r}
#| echo: false
source("../plot/theme_jjf.R")
```

```{python}
#| echo: false
exec(open("../helper-levels.py").read())
```

```{r}
#| echo: false
palette <- c("black", palette_jjf(length(reticulate::py$factors)))
names(palette) <- c("Overall", reticulate::py$factors)
```

```{python}
#| echo: false
overlap_df = overlap_df.dropna()
```

# Decomposition

Underlying returns are structural bets that can be analyzed through dimension reduction techniques such as principal component analysis (PCA). Most empirical studies apply PCA to a covariance matrix (*note: for multi-asset portfolios, the correlation matrix is often used instead because asset-class variances are on different scales*) of equity returns (yield changes) and find that movements in the equity markets (yield curve) can be explained by a subset of principal components. For example, the first three principal components of the yield curve can be interpreted as shift, twist, and butterfly, respectively.

$$
\begin{aligned}
\boldsymbol{\Sigma}&=\lambda_{1}\mathbf{v}_{1}\mathbf{v}_{1}^\mathrm{T}+\lambda_{2}\mathbf{v}_{2}\mathbf{v}_{2}^\mathrm{T}+\cdots+\lambda_{k}\mathbf{v}_{k}\mathbf{v}_{k}^\mathrm{T}\\
&=V\Lambda V^\mathrm{T}
\end{aligned}
$$

-   <https://www.r-bloggers.com/fixing-non-positive-definite-correlation-matrices-using-r-2/>

```{python}
def eigen(x):
  
  L, V = np.linalg.eigh(np.cov(x.T, ddof = 1))
  L = L[::-1]
  V = V[:, ::-1]
  
  result = {
    "values": L,
    "vectors": V
  }
    
  return result
```

```{python}
def eigen_decomp(x, comps):
  
  LV = eigen(x)
  L = LV["values"][:comps]
  V = LV["vectors"][:, :comps]
  
  result = np.dot(V, np.multiply(L[:, None], V.T))
  
  return result
```

```{python}
comps = 1
```

```{python}
eigen_decomp(overlap_df, comps) * scale["periods"] * scale["overlap"]
```

```{python}
# np.cov(overlap_df.T) * scale["periods"] * scale["overlap"]
```

# Variance

We often look at the proportion of variance explained by the first $i$ principal components as an indication of how many components are needed.

$$
\begin{aligned}
\frac{\sum_{j=1}^{i}{\lambda_{j}}}{\sum_{j=1}^{k}{\lambda_{j}}}
\end{aligned}
$$

```{python}
def variance_explained(x):
  
  LV = eigen(x)
  L = LV["values"]
  
  result = L.cumsum() / L.sum()
  
  return result
```

```{python}
variance_explained(overlap_df)
```

# Similarity

Also, a challenge of rolling PCA is to try to match the eigenvectors: may need to change the sign and order.

$$
\begin{aligned}
\text{similarity}=\frac{\mathbf{v}_{t}\cdot\mathbf{v}_{t-1}}{\|\mathbf{v}_{t}\|\|\mathbf{v}_{t-1}\|}
\end{aligned}
$$

```{python}
def similarity(V, V0):
  
  n_cols_v = V.shape[1]
  n_cols_v0 = V0.shape[1]
  result = np.zeros((n_cols_v, n_cols_v0))
  
  for i in range(n_cols_v):
    for j in range(n_cols_v0):
      result[i, j] = np.dot(V[:, i], V0[:, j]) / \
        np.sqrt(np.dot(V[:, i], V[:, i]) * np.dot(V0[:, j], V0[:, j]))
      
  return result
```

```{python}
def roll_eigen1(x, width, comp):
    
  n_rows = len(x)
  result_ls = []
  
  for i in range(width - 1, n_rows):
      
    idx = range(max(i - width + 1, 0), i + 1)
    
    LV = eigen(x.iloc[idx])
    V = LV["vectors"]
    
    result_ls.append(V[:, comp - 1])
  
  result_df = pd.DataFrame(result_ls, index = x.index[(width - 1):],
                           columns = x.columns)
  
  return result_df  
```

```{python}
comp = 1
```

```{python}
raw_df = roll_eigen1(overlap_df, width, comp)
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
raw_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$raw_df, keep.rownames = "index"), id.vars = "index")
raw_mlt[ , index := as.Date(index)]
raw_plt <- plot_ts(raw_mlt, title = "Eigenvector 1Y (raw)")
print(raw_plt)
```

-   <https://quant.stackexchange.com/a/3095>
-   <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=4400158>

```{python}
def roll_eigen2(x, width, comp):
    
  n_rows = len(x)
  V_ls = []
  result_ls = []
  
  for i in range(width - 1, n_rows):
      
    idx = range(max(i - width + 1, 0), i + 1)
    
    LV = eigen(x.iloc[idx])
    V = LV["vectors"]
    
    if i > width - 1:
        
      # cosine = np.dot(V.T, V_ls[-1])
      cosine = similarity(V, V_ls[-1])
      order = np.argmax(np.abs(cosine), axis = 0) # assumes a one-to-one match between windows
      V = np.sign(np.diag(cosine[order, :])) * V[:, order]

    V_ls.append(V)
    result_ls.append(V[:, comp - 1])
  
  result_df = pd.DataFrame(result_ls, index = x.index[(width - 1):],
                           columns = x.columns)
  
  return result_df
```

```{python}
clean_df = roll_eigen2(overlap_df, width, comp)
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
clean_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$clean_df, keep.rownames = "index"), id.vars = "index")
clean_mlt[ , index := as.Date(index)]
clean_plt <- plot_ts(clean_mlt, title = "Eigenvector 1Y (clean)")
print(clean_plt)
```

# Implied shocks

Implied shocks are the product of the $n$th eigenvector and the square root of the $n$th eigenvalue (annualized using the square root of time):

$$
\begin{aligned}
\sqrt{\lambda_{n}}\mathbf{v}_{n}
\end{aligned}
$$

```{python}
def roll_shocks(x, width, comp):
    
  n_rows = len(x)
  V_ls = []
  result_ls = []
  
  for i in range(width - 1, n_rows):
    
    idx = range(max(i - width + 1, 0), i + 1)
    
    LV = eigen(x.iloc[idx])
    L = LV["values"]
    V = LV["vectors"]
    
    if i > width - 1:

      # cosine = np.dot(V.T, V_ls[-1])
      cosine = similarity(V, V_ls[-1])
      order = np.argmax(np.abs(cosine), axis = 0) # assumes a one-to-one match between windows
      L = L[order]
      V = np.sign(np.diag(cosine[order, :])) * V[:, order]
    
    shocks = np.sqrt(L[comp - 1]) * V[:, comp - 1]
    V_ls.append(V)
    result_ls.append(shocks)
    
  result_df = pd.DataFrame(result_ls, index = x.index[(width - 1):],
                           columns = x.columns)
  
  return result_df
```

```{python}
shocks_df = roll_shocks(overlap_df, width, comp) * np.sqrt(scale["periods"] * scale["overlap"])
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
shocks_mlt <- data.table::melt(data.table::as.data.table(reticulate::py$shocks_df, keep.rownames = "index"), id.vars = "index")
shocks_mlt[ , index := as.Date(index)]
shocks_plt <- plot_ts(shocks_mlt, title = "Implied shocks 1Y using PC1 (%)", multiple = 100)
print(shocks_plt)
```
