---
title: "Eigen"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - r
execute:
  freeze: false
editor: 
  mode: source
---

```{r}
factors_r <- c("SP500", "DTWEXAFEGS") # "SP500" does not contain dividends; note: "DTWEXM" discontinued as of Jan 2020
factors_d <- c("DGS10", "BAMLH0A0HYM2")
```

```{r}
#| echo: false
#| output: false # Registered S3 method overwritten by 'quantmod'
source("../plot/theme_jjf.R")
source("../helper-levels.R")
```

```{r}
#| echo: false
palette <- c("black", palette_jjf(length(factors)))
names(palette) <- c("Overall", factors)
```

```{r}
#| echo: false
overlap_xts <- na.omit(overlap_xts)
```

# Decomposition

Underlying returns are structural bets that can be analyzed through dimension reduction techniques such as principal components analysis (PCA). Most empirical studies apply PCA to a covariance matrix (*note: for multi-asset portfolios, use the correlation matrix because asset-class variances are on different scales*) of equity returns (yield changes) and find that movements in the equity markets (yield curve) can be explained by a subset of principal components. For example, the yield curve can be decomposed in terms of shift, twist, and butterfly, respectively.

$$
\begin{aligned}
\boldsymbol{\Sigma}&=\lambda_{1}\mathbf{v}_{1}\mathbf{v}_{1}^\mathrm{T}+\lambda_{2}\mathbf{v}_{2}\mathbf{v}_{2}^\mathrm{T}+\cdots+\lambda_{k}\mathbf{v}_{k}\mathbf{v}_{k}^\mathrm{T}\\
&=V\Lambda V^{\mathrm{T}}
\end{aligned}
$$

-   <https://www.r-bloggers.com/fixing-non-positive-definite-correlation-matrices-using-r-2/>

```{r}
eigen_decomp <- function(x, comps) {
  
  LV <- eigen(cov(x))
  L <- LV[["values"]][1:comps]
  V <- LV[["vectors"]][ , 1:comps]
  
  result <- V %*% sweep(t(V), 1, L, "*")
  
  return(result)
    
}
```

```{r}
comps <- 1
```

```{r}
eigen_decomp(overlap_xts, comps) * scale[["periods"]] * scale[["overlap"]]
```

```{r}
# cov(overlap_xts) * scale[["periods"]] * scale[["overlap"]]
```

# Variance

We often look at the proportion of variance explained by the first $i$ principal components as an indication of how many components are needed.

$$
\begin{aligned}
\frac{\sum_{j=1}^{i}{\lambda_{j}}}{\sum_{j=1}^{k}{\lambda_{j}}}
\end{aligned}
$$

```{r}
variance_explained <- function(x) {
    
  LV <- eigen(cov(x))
  L <- LV[["values"]]
  
  result <- cumsum(L) / sum(L)
  
  return(result)
    
}
```

```{r}
variance_explained(overlap_xts)
```

# Similarity

Also, a challenge of rolling PCA is to try to match the eigenvectors: may need to change the sign and order.

$$
\begin{aligned}
\text{similarity}=\frac{\mathbf{v}_{t}\cdot\mathbf{v}_{t-1}}{\|\mathbf{v}_{t}\|\|\mathbf{v}_{t-1}\|}
\end{aligned}
$$

```{r}
similarity <- function(V, V0) {
    
  n_cols_v <- ncol(V)
  n_cols_v0 <- ncol(V0)
  result <- matrix(0, nrow = n_cols_v, ncol = n_cols_v0)
  
  for (i in 1:n_cols_v) {
    for (j in 1:n_cols_v0) {
      result[i, j] <- crossprod(V[ , i], V0[ , j]) /
        sqrt(crossprod(V[ , i]) * crossprod(V0[ , j]))
    }
  }
  
  return(result)
    
}
```

```{r}
roll_eigen1 <- function(x, width, comp) {
    
  n_rows <- nrow(x)
  result_ls <- list()
  
  for (i in width:n_rows) {
    
    idx <- max(i - width + 1, 1):i
    
    LV <- eigen(cov(x[idx, ]))
    V <- LV[["vectors"]]
    
    result_ls <- append(result_ls, list(V[ , comp]))
    
  }
  
  result <- do.call(rbind, result_ls)
  result <- xts::xts(result, zoo::index(x)[width:n_rows])
  colnames(result) <- colnames(x)
  
  return(result)
    
}
```

```{r}
comp <- 1
```

```{r}
raw_df <- roll_eigen1(overlap_xts, width, comp)
```

```{r}
# # install.packages("devtools")
# devtools::install_github("jasonjfoster/rolleigen") # roll (>= 1.1.7)
# raw_df <- rolleigen::roll_eigen(overlap_xts, width, order = TRUE)[["vectors"]][ , comp, ]
# raw_df <- xts::xts(t(raw_df), zoo::index(overlap_xts))
# colnames(raw_df) <- colnames(overlap_xts)
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
raw_mlt <- data.table::melt(data.table::as.data.table(raw_df), id.vars = "index")
raw_plt <- plot_ts(raw_mlt, title = "Eigenvector 1Y")
print(raw_plt)
```

-   <https://quant.stackexchange.com/a/3095>
-   <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=4400158>

```{r}
roll_eigen2 <- function(x, width, comp) {
    
  n_rows <- nrow(x)
  V_ls <- list()
  result_ls <- list()
  
  for (i in width:n_rows) {
    
    idx <- max(i - width + 1, 1):i
    
    LV <- eigen(cov(x[idx, ]))
    V <- LV[["vectors"]]
    
    if (i > width) {
      
      # cosine <- crossprod(V, V_ls[[length(V_ls)]])
      cosine <- similarity(V, V_ls[[length(V_ls)]])
      order <- apply(abs(cosine), 1, which.max)
      V <- t(sign(diag(cosine[ , order])) * t(V[ , order]))
      
    }
    
    V_ls <- append(V_ls, list(V))
    result_ls <- append(result_ls, list(V[ , comp]))
    
  }
  
  result <- do.call(rbind, result_ls)
  result <- xts::xts(result, zoo::index(x)[width:n_rows])
  colnames(result) <- colnames(x)
  
  return(result)
    
}
```

```{r}
clean_df <- roll_eigen2(overlap_xts, width, comp)
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
clean_mlt <- data.table::melt(data.table::as.data.table(clean_df), id.vars = "index")
clean_plt <- plot_ts(clean_mlt, title = "Eigenvector 1Y")
print(clean_plt)
```

# Implied shocks

Product of the $n$th eigenvector and square root of the $n$th eigenvalue:

```{r}
roll_shocks <- function(x, width, comp) {
  
  n_rows <- nrow(x)
  V_ls <- list()
  result_ls <- list()
  
  for (i in width:n_rows) {
    
    idx <- max(i - width + 1, 1):i
    
    LV <- eigen(cov(x[idx, ]))
    L <- LV[["values"]]
    V <- LV[["vectors"]]
    
    if (length(V_ls) > 1) {
      
      # cosine <- crossprod(V, V_ls[[length(V_ls)]])
      cosine <- similarity(V, V_ls[[length(V_ls)]])
      order <- apply(abs(cosine), 1, which.max)
      L <- L[order]
      V <- t(sign(diag(cosine[ , order])) * t(V[ , order]))
      
    }
    
    shocks <- sqrt(L[comp]) * V[ , comp]
    V_ls <- append(V_ls, list(V))
    result_ls <- append(result_ls, list(shocks))
    
  }
  
  result <- do.call(rbind, result_ls)
  result <- xts::xts(result, zoo::index(x)[width:n_rows])
  colnames(result) <- colnames(x)
  
  return(result)
    
}
```

```{r}
shocks_xts <- roll_shocks(overlap_xts, width, comp) * sqrt(scale[["periods"]] * scale[["overlap"]])
```

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
shocks_mlt <- data.table::melt(data.table::as.data.table(shocks_xts), id.vars = "index")
shocks_plt <- plot_ts(shocks_mlt, title = "Implied shocks 1Y using PC1 (%)", multiple = 100)
print(shocks_plt)
```
