---
title: "Risk"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - r
execute:
  freeze: false
editor:
  mode: source
---

```{r}
factors_r <- c("SP500", "DTWEXAFEGS") # "SP500" does not contain dividends; note: "DTWEXM" discontinued as of Jan 2020
factors_d <- c("DGS10", "BAMLH0A0HYM2")
```

```{r}
tickers <- c("BAICX") # fund inception date is "2011-11-28"
```

```{r}
intercept <- TRUE
```

```{r}
#| echo: false
#| output: false # Registered S3 method overwritten by 'quantmod': method as.zoo.data.frame from zoo
source("../helper-levels.R")
source("../helper-prices.R")
# weights <- 0.9 ^ ((width - 1):0)
weights <- rep(1, width)

overlap_xts <- roll::roll_mean(returns_xts, scale[["overlap"]], min_obs = 1)

overlap_xts <- tail(na.omit(overlap_xts), width) # same dimension as `weights`
overlap_x_xts <- overlap_xts[ , factors]
overlap_y_xts <- overlap_xts[ , tickers]
```

# Regression analysis

A factor model expresses fund returns as exposures to a set of market factors, which can be estimated with regression analysis. To measure the risk of a fund, estimate factor exposures for an allocation fund, decompose volatility into factor contributions, and then stress the factors under scenarios.

The example uses equity, currency, interest rate, and credit factors with a trailing one-year window of overlapping returns and equal weights (*note: rates and spreads are negated yield changes, i.e. duration-style returns, and volatility is annualized using the square root of time*).

## Ordinary least squares

### Coefficients

Solve for the weighted least squares coefficients in matrix form:

$$
\begin{aligned}
\hat{\beta}=(X^\mathrm{T}WX)^{-1}X^\mathrm{T}Wy
\end{aligned}
$$

-   <https://faculty.washington.edu/ezivot/research/factormodellecture_handout.pdf>

```{r}
lm_coef <- function(x, y, weights, intercept) {
  
  # cbind.xts() changes names, e.g., "(Intercept)" => "X.Intercept."
  if (intercept) x <- cbind("(Intercept)" = 1, as.matrix(x))
  
  result <- solve(crossprod(x, diag(weights)) %*% x) %*% crossprod(x, diag(weights) %*% y)
  
  return(result)
  
}
```

```{r}
t(lm_coef(overlap_x_xts, overlap_y_xts, weights, intercept))
```

```{r}
if (intercept) {
  form <- reformulate(termlabels = factors, response = tickers)
} else {
  form <- reformulate(termlabels = factors, response = tickers, intercept = FALSE)
}

fit <- lm(form, data = overlap_xts, weights = weights)
    
coef(fit)
```

### R-squared

Then compute the weighted R-squared, i.e. the proportion of variance explained by the factors:

$$
\begin{aligned}
R^{2}=\frac{\hat{\beta}^\mathrm{T}(X^\mathrm{T}WX)\hat{\beta}}{y^\mathrm{T}Wy}
\end{aligned}
$$

```{r}
lm_rsq <- function(x, y, weights, intercept) {
        
  coef <- lm_coef(x, y, weights, intercept)
  
  if (intercept) {
    
    x <- cbind(1, x)
    x <- sweep(x, 2, apply(x, 2, weighted.mean, w = weights), "-")
    y <- sweep(y, 2, apply(y, 2, weighted.mean, w = weights), "-")
      
  }
  
  result <- (t(coef) %*% (crossprod(x, diag(weights)) %*% x) %*% coef) / (crossprod(y, diag(weights)) %*% y)
  colnames(result) <- "R-squared"
  
  return(result)
    
}
```

```{r}
lm_rsq(overlap_x_xts, overlap_y_xts, weights, intercept)
```

```{r}
summary(fit)$r.squared
```

### Standard errors

Also compute the standard errors of the intercept and coefficients using the residual variance:

$$
\begin{aligned}
\sigma_{\hat{\beta}}^{2}&=\sigma_{\varepsilon}^{2}(X^\mathrm{T}WX)^{-1}\\
&=\frac{(1-R^{2})(y^\mathrm{T}Wy)}{n-p}(X^\mathrm{T}WX)^{-1}\\
&=\frac{SSE}{df_{E}}(X^\mathrm{T}WX)^{-1}\\
\sigma_{\hat{\alpha}}^{2}&=\sigma_{\varepsilon}^{2}\left(\frac{1}{\sum_{i=1}^{n}w_{i}}+\mu^\mathrm{T}(X^\mathrm{T}WX)^{-1}\mu\right)
\end{aligned}
$$

-   <http://people.duke.edu/~rnau/mathreg.htm>

```{r}
lm_se <- function(x, y, weights, intercept) {
    
  n_rows <- nrow(x)
  n_cols <- ncol(x)
  
  rsq <- lm_rsq(x, y, weights, intercept)
  
  if (intercept) {
    
    # cbind.xts() changes names, e.g., "(Intercept)" => "X.Intercept."
    x <- cbind("(Intercept)" = 1, as.matrix(x))
    y <- sweep(y, 2, apply(y, 2, weighted.mean, w = weights), "-")
    
    df_resid <- n_rows - n_cols - 1
      
  } else {
    df_resid <- n_rows - n_cols
  }
  
  var_y <- crossprod(y, diag(weights)) %*% y
  var_resid <- as.numeric((1 - rsq) * var_y / df_resid)
  
  result <- sqrt(var_resid * diag(solve(crossprod(x, diag(weights)) %*% x)))
  
  return(result)
    
}
```

```{r}
lm_se(overlap_x_xts, overlap_y_xts, weights, intercept)
```

```{r}
coef(summary(fit))[ , "Std. Error"]
```

### Shapley values

Since the factors are correlated, allocate the R-squared to each factor using the Shapley value, i.e. the average marginal contribution across all combinations of factors:

$$
R^{2}_{i}=\sum_{S\subseteq N\setminus\{i\}}{\frac{|S|!\;(n-|S|-1)!}{n!}}(R^{2}(S\cup\{i\})-R^{2}(S))
$$

-   <https://real-statistics.com/multiple-regression/shapley-owen-decomposition/>

```{r}
lm_shap <- function(x, y, weights, intercept) {
  
  n_rows <- nrow(x)
  n_cols <- ncol(x)
  n_combn <- 2 ^ n_cols
  n_vec <- array(0, n_combn)
  ix_mat <- matrix(0, nrow = n_cols, ncol = n_combn)
  rsq <- array(0, n_combn)
  result <- array(0, n_cols)
  
  # number of binary combinations
  for (k in 1:n_combn) {
    
    n <- 0
    n_size <- k - 1
    
    # find the binary combination
    for (j in 1:n_cols) {
      
      if (n_size %% 2 == 0) {
        
        n <- n + 1
        
        ix_mat[j, k] <- j
        
      }
      
      n_size <- n_size %/% 2
      
    }
    
    n_vec[k] <- n
    
    if (n > 0) {
      
      ix_subset <- which(ix_mat[ , k] != 0)
      x_subset <- x[ , ix_subset]
      
      rsq[k] <- lm_rsq(x_subset, y, weights, intercept)

    }
    
  }

  # calculate the exact Shapley value for r-squared
  for (j in 1:n_cols) {

    ix_pos <- which(ix_mat[j, ] != 0)
    ix_neg <- which(ix_mat[j, ] == 0)
    ix_n <- n_vec[ix_neg]
    rsq_diff <- rsq[ix_pos] - rsq[ix_neg]

    for (k in 1:(n_combn / 2)) {

      s <- ix_n[k]
      weight <- factorial(s) * factorial(n_cols - s - 1) / factorial(n_cols)
      result[j] <- result[j] + weight * rsq_diff[k]

    }

  }

  names(result) <- colnames(x)

  return(result)
  
}
```

```{r}
lm_shap(overlap_x_xts, overlap_y_xts, weights, intercept)
```

## Principal component regression

Alternatively, to address correlated factors, regress on the principal components of the factors and then map the coefficients to the original variables:

```{r}
#| output: false # Attaching package: 'pls' - The following object is masked from 'package:stats': loadings
library(pls) # "Error in mvrValstats(object = fit, estimate = 'train'): could not find function 'mvrValstats'"
```

```{r}
comps <- 1
```

### Coefficients

$$
\begin{aligned}
W_{k}&=\mathbf{X}V_{k}=[\mathbf{X}\mathbf{v}_{1},\ldots,\mathbf{X}\mathbf{v}_{k}]\\
{\widehat{\gamma}}_{k}&=\left(W_{k}^\mathrm{T}W_{k}\right)^{-1}W_{k}^\mathrm{T}\mathbf{Y}\\
{\widehat{\boldsymbol{\beta}}}_{k}&=V_{k}{\widehat{\gamma}}_{k}
\end{aligned}
$$

-   <https://en.wikipedia.org/wiki/Principal_component_regression>

```{r}
pcr_coef <- function(x, y, comps) {
  
  x <- sweep(x, 2, colMeans(x), "-")
  LV <- eigen(cov(x))
  V <- LV[["vectors"]]
  
  W <- x %*% V
  gamma <- solve(crossprod(W)) %*% (crossprod(W, y))
  
  result <- V[ , 1:comps] %*% as.matrix(gamma[1:comps])
  
  return(result)
  
}
```

```{r}
scale_x_xts <- scale(overlap_x_xts)
```

```{r}
t(pcr_coef(scale_x_xts, overlap_y_xts, comps))
```

```{r}
t(pcr_coef(overlap_x_xts, overlap_y_xts, comps))
```

```{r}
fit <- pls::pcr(reformulate(termlabels = ".", response = tickers),
                data = merge(scale_x_xts, overlap_y_xts), ncomp = comps)
coef(fit)[ , , 1]
```

### R-squared

```{r}
pcr_rsq <- function(x, y, comps) {
  
  coef <- pcr_coef(x, y, comps)
  
  x <- sweep(x, 2, colMeans(x), "-")
  y <- sweep(y, 2, colMeans(y), "-")
  
  result <- (t(coef) %*% crossprod(x) %*% coef) / crossprod(y)
  colnames(result) <- "R-squared"
  
  return(result)
  
}
```

```{r}
pcr_rsq(scale_x_xts, overlap_y_xts, comps)
```

```{r}
pcr_rsq(overlap_x_xts, overlap_y_xts, comps)
```

```{r}
pls::R2(fit)$val[comps + 1]
```

### Standard errors

$$
\begin{aligned}
\text{Var}({\widehat{\boldsymbol{\beta}}}_{k})&=\sigma^{2}V_{k}(W_{k}^\mathrm{T}W_{k})^{-1}V_{k}^\mathrm{T}\\
&=\sigma^{2}V_{k}\text{diag}\left(\lambda_{1}^{-1},\ldots,\lambda_{k}^{-1}\right)V_{k}^\mathrm{T}\\
&=\sigma^{2}\sum_{j=1}^{k}{\frac{\mathbf{v}_{j}\mathbf{v}_{j}^\mathrm{T}}{\lambda_{j}}}
\end{aligned}
$$

Standard errors are equivalent to the standard errors from a regression on the principal component scores mapped to the original coefficients, i.e. use the residual degrees of freedom for the $k$ retained components:

```{r}
pcr_se <- function(x, y, comps) {

  n_rows <- nrow(x)

  rsq <- pcr_rsq(x, y, comps)

  y <- sweep(y, 2, colMeans(y), "-")

  df_resid <- n_rows - comps - 1
  
  var_y <- crossprod(y)
  var_resid <- as.numeric((1 - rsq) * var_y / df_resid)
  
  LV <- eigen(cov(x))
  L <- LV$values[1:comps] * (n_rows - 1)
  V <- LV$vectors[ , 1:comps]
  
  result <- sqrt(var_resid * diag(V %*% sweep(t(V), 1, 1 / L, "*")))
  
  return(result)
  
}
```

```{r}
pcr_se(scale_x_xts, overlap_y_xts, comps)
```

```{r}
pcr_se(overlap_x_xts, overlap_y_xts, comps)
```

```{r}
LV <- eigen(cov(scale_x_xts))
V <- as.matrix(LV[["vectors"]][ , 1:comps])
W <- scale_x_xts %*% V

fit <- lm(zoo::coredata(overlap_y_xts) ~ W)
sqrt(diag(V %*% as.matrix(vcov(fit)[-1, -1]) %*% t(V)))
```

## Partial least squares

# Risk decomposition

Next, decompose the volatility of the fund using the factor model:

## Standalone risk

Standalone risk is the volatility of each factor exposure in isolation, i.e. ignoring correlations, so the components do not sum to the total:

$$
\begin{aligned}
\text{SAR}_{k}&=\sqrt{w_{k}^{2}\sigma_{k}^{2}}\\
\text{SAR}_{\varepsilon}&=\sqrt{(1-R^{2})\sigma_{y}^{2}}
\end{aligned}
$$

```{r}
lm_sar <- function(x, y, weights, intercept) {
  
  coef <- lm_coef(x, y, weights, intercept)
  rsq <- lm_rsq(x, y, weights, intercept)
  
  if (intercept) x <- cbind(1, x)
  
  sigma <- cov.wt(cbind(x, y), wt = weights, center = intercept)$cov
  sar <- coef ^ 2 * diag(sigma[-ncol(sigma), -ncol(sigma)])
  sar_eps <- (1 - rsq) * sigma[ncol(sigma), ncol(sigma)]
  
  result <- sqrt(c(sigma[ncol(sigma), ncol(sigma)],
                   sar,
                   sar_eps))
  names(result) <- c("total", rownames(coef), "eps")
  
  return(result)
  
}
```

```{r}
lm_sar(overlap_x_xts, overlap_y_xts, weights, intercept) * sqrt(scale[["periods"]] * scale[["overlap"]])
```

## Risk contribution

Marginal contribution to risk allocates the total volatility to each factor exposure, i.e. including correlations, so the components sum to the total:

$$
\begin{aligned}
\text{MCR}_{k}&=w_{k}\frac{\partial\sigma_{y}}{\partial w_{k}}\\
&=w_{k}\frac{(\Sigma w)_{k}}{\sigma_{y}}\\
\text{MCR}_{\varepsilon}&=\sigma_{y}-\sum_{k=1}^{n}\text{MCR}_{k}
\end{aligned}
$$

-   <https://bookdown.org/compfinezbook/introcompfinr/Portfolio-risk-reports.html>

```{r}
lm_mcr <- function(x, y, weights, intercept) {
  
  coef <- lm_coef(x, y, weights, intercept)
  rsq <- lm_rsq(x, y, weights, intercept)
  
  if (intercept) x <- cbind(1, x)
  
  sigma <- cov.wt(cbind(x, y), wt = weights, center = intercept)$cov
  mcr <- coef * sigma[-ncol(sigma), -ncol(sigma)] %*% coef / sqrt(sigma[ncol(sigma), ncol(sigma)])
  mcr_eps <- sqrt(sigma[ncol(sigma), ncol(sigma)]) - sum(mcr)
  
  result <- c(sqrt(sigma[ncol(sigma), ncol(sigma)]),
              mcr,
              mcr_eps)
  names(result) <- c("total", rownames(coef), "eps")
  
  return(result)
  
}
```

```{r}
lm_mcr(overlap_x_xts, overlap_y_xts, weights, intercept) * sqrt(scale[["periods"]] * scale[["overlap"]])
```

# Scenario analysis

Finally, estimate the impact of scenarios on the fund using the factor exposures:

## Implied shocks

Implied shocks translate a scenario for a subset of factors into consistent shocks for all factors using weighted least squares:

$$
\begin{aligned}
\hat{\beta}&=(Z^\mathrm{T}WZ)^{-1}Z^\mathrm{T}WX
\end{aligned}
$$

```{r}
implied_shocks <- function(shocks, x, z, weights) {
  
  beta <- solve(crossprod(z, diag(weights) %*% z)) %*% crossprod(z, diag(weights) %*% x)
  
  result <- shocks %*% beta
  
  return(result)
  
}
```

```{r}
shocks <- c(-0.1, 0.1)
overlap_z_xts <- overlap_x_xts[ , 1:2]
```

```{r}
implied_shocks(shocks, overlap_x_xts, overlap_z_xts, weights)
```

## Stress P&L

Stress P&L applies the implied shocks to the factor exposures. Note: if an intercept is included, then alpha is stressed along with the factors, i.e. the constant is passed through the implied shocks regression:

```{r}
pnl_stress <- function(shocks, x, y, z, weights, intercept) {
  
  coef <- lm_coef(x, y, weights, intercept)
  
  # cbind.xts() changes names, e.g., "(Intercept)" => "X.Intercept."
  if (intercept) x <- cbind("(Intercept)" = 1, as.matrix(x))
  
  result <- t(coef) * implied_shocks(shocks, x, z, weights)
  
  return(result)    
  
}
```

```{r}
pnl_stress(shocks, overlap_x_xts, overlap_y_xts, overlap_z_xts, weights, intercept)
```