---
title: "Alpha"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - r
draft: true
execute:
  freeze: false
editor: 
  mode: source
---

```{r}
#| echo: false
source("../plot/theme_jjf.R")
```

# Question

Could an optimally combined portfolio of hedge funds have won Warren Buffet's 10-year bet against the S&P 500?

-   <https://longbets.org/362/>

```{r}
load("returns.rda") # fund-of-funds and S&P 500 index fund returns for the 10-year bet
```

# Optimization

```{r}
# install.packages("pak")
# pak::pak("jasonjfoster/rolloptim")
```

```{r}
#| output: false # Attaching package: 'CVXR' - The following object is masked from 'package:stats': power
# library(CVXR)
```

```{r}
# min_rss_optim <- function(x, y) {
# 
#   n_rows <- nrow(x)
#   x <- as.matrix(x)
#   y <- as.numeric(y)
#   params <- Variable(ncol(x))
#   
#   obj <- Minimize(sum_squares(y - x %*% params))
#   cons <- list(sum(params) == 1, params >= 0, params <= 1)
#   prob <- Problem(obj, cons)
#   
#   psolve(prob)
#
#   result <- value(params)
#   
#   return(result)
# 
# }
```

# Bootstrapping

```{r}
# library(NNS)
```

```{r}
boot <- function(n, p = 1) {

  idx <- sample(1:n, 1)
  idx_ls <- list(idx)
  
  for (i in 2:n) {
    if (runif(1) < 1 / p) {
      idx <- sample(1:n, 1)
    } else {

      idx <- idx + 1

      if (idx > n) {
        idx <- 1
      }

    }
    
    idx_ls <- append(idx_ls, list(idx))

  }

  result <- do.call(rbind, idx_ls)
  
  return(result)

}
```

```{r}
set.seed(5640)
n_rows <- nrow(returns)
n_cols <- ncol(returns)
n_boot <- 1000
optim_weights_ls <- list()
optim_alpha_ls <- list()
optim_te_ls <- list()
# returns_boot <- array(NA, dim = c(n_rows, n_cols, n_boot)) # time, asset, and bootstrap
```

```{r}
# for (j in 1:n_cols) {
# 
#   col_boot <- NNS.meboot(as.vector(returns[ , j]), reps = n_boot,
#                          rho = 1, type = "pearson")["replicates", ]$replicates
# 
#   returns_boot[ , j, ] <- col_boot
# 
# }
```

```{r}
# colnames(returns_boot) <- colnames(returns)
```

```{r}
for (z in 1:n_boot) {
  
  # idx <- sample(1:n_rows, n_rows, replace = TRUE)
  idx <- boot(n_rows, p = 2)
  # boot_sample <- returns_boot[ , , z]
  
  x_boot <- returns[idx, colnames(returns) != "S&P Index Fund"]
  y_boot <- returns[idx, "S&P Index Fund"]
  # x_boot <- boot_sample[ , colnames(boot_sample) != "S&P Index Fund"]
  # y_boot <- boot_sample[ , "S&P Index Fund"]
  
  # optim_weights <- min_rss_optim(x_boot, y_boot)
  
  xx <- roll::roll_crossprod(x_boot, x_boot, width = nrow(x_boot))
  xy <- roll::roll_crossprod(x_boot, y_boot, width = nrow(x_boot))
  
  optim_weights <- rolloptim::roll_min_rss(xx, xy, total = 1, lower = 0, upper = 1)[nrow(x_boot), ]
  
  optim_returns <- x_boot %*% optim_weights
  optim_port <- prod(1 + optim_returns) ^ (1 / n_rows) - 1
  optim_bench <- prod(1 + y_boot) ^ (1 / n_rows) - 1
  optim_alpha <- optim_port - optim_bench
  optim_te <- sd(optim_returns - y_boot)
  
  optim_weights_ls <- append(optim_weights_ls, list(t(optim_weights)))
  optim_alpha_ls <- append(optim_alpha_ls, list(optim_alpha))
  optim_te_ls <- append(optim_te_ls, list(optim_te))

}
```

```{r}
optim_weights <- do.call(rbind, optim_weights_ls)
optim_alpha <- do.call(rbind, optim_alpha_ls)
optim_te <- do.call(rbind, optim_te_ls)
```

# Answer

Analysis shows that `{r} round(mean(optim_alpha > 0), 2) * 100`% of the optimized portfolios produced alpha values greater than zero, which means that `{r} round(mean(optim_alpha <= 0), 2) * 100`% of optimized combinations did not outperform the S&P 500. Also the median information ratio across the optimized portfolios, i.e. alpha per unit of tracking error, is `{r} round(median(optim_alpha / optim_te), 2)`.

```{r}
#| echo: false
#| fig.width: 6
#| fig.height: 3
alpha_dt <- data.table::data.table(optim_alpha = as.vector(optim_alpha))
q_vals <- quantile(alpha_dt$optim_alpha, probs = c(0.05, 0.50, 0.95))

hist_data <- hist(alpha_dt$optim_alpha, breaks = 30, plot = FALSE)
bin_width <- diff(hist_data$breaks)[1]
hist_dt <- data.table::data.table(
  mids = hist_data$mids * 100,
  freq_pct = hist_data$counts / sum(hist_data$counts) * 100
)

dens <- density(alpha_dt$optim_alpha)
density_dt <- data.table::data.table(
  alpha_pct = dens$x * 100,
  density = dens$y * bin_width * 100 # convert density to frequency (%)
)

y_max <- max(hist_dt$freq_pct)

alpha_plt <- ggplot2::ggplot() +
  theme_jjf() +
  ggplot2::labs(title = "Alpha distribution", x = "Alpha (%)", y = "Frequency (%)") +
  ggplot2::geom_col(
    data = hist_dt,
    ggplot2::aes(x = mids, y = freq_pct),
    width = bin_width * 100
  ) +
  ggplot2::geom_line(
    data = density_dt,
    ggplot2::aes(x = alpha_pct, y = density),
    color = "black"
  ) +
  ggplot2::geom_vline(xintercept = 0, color = "black") +
  ggplot2::geom_vline(xintercept = q_vals * 100, linetype = "dashed", color = palette_jjf(1)) +
  ggplot2::annotate("text", x = q_vals[1] * 100, y = y_max, label = "5%",
                    color = palette_jjf(1), hjust = 1) +
  ggplot2::annotate("text", x = q_vals[2] * 100, y = y_max, label = "Median",
                    color = palette_jjf(1), hjust = 1) +
  ggplot2::annotate("text", x = q_vals[3] * 100, y = y_max, label = "95%",
                    color = palette_jjf(1), hjust = 1)
print(alpha_plt)
```