---
title: "Securities"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - analysis
  - finance
  - r
---

```{r, echo = FALSE, message = FALSE}
source("../plot/theme_jjf.R")
```

```{r, message = FALSE}
library(quantmod)
library(roll)
library(data.table)
```

```{r, echo = FALSE}
options("getSymbols.warning4.0" = FALSE)
```

```{r}
factors_r <- c("SP500", "DTWEXAFEGS") # "SP500" does not contain dividends; note: "DTWEXM" discontinued as of Jan 2020
factors_d <- c("DGS10", "BAMLH0A0HYM2")
factors <- c(factors_r, factors_d)
width <- 252
scale <- list("periods" = 252, "overlap" = 5)
```

```{r, echo = FALSE}
palette <- c("black", palette_jjf(length(factors)))
names(palette) <- c("Overall", factors)
```

```{r, results = "hide"}
getSymbols(factors, src = "FRED")
levels_xts <- do.call(merge, c(lapply(factors, function(i) get(i)), all = TRUE))
```

```{r}
returns_xts <- do.call(merge, lapply(factors, function(i) {
    if (i %in% factors_r) {
        diff(log((levels_xts[ , i])))
    } else if (i %in% factors_d) {
        -diff(levels_xts[ , i]) / 100
    }    
}))
overlap_xts <- roll_mean(returns_xts, scale[["overlap"]], min_obs = 1, na_restore = TRUE)
```

```{r}
sd_xts <- roll_sd(overlap_xts, width, min_obs = 1) * sqrt(scale[["periods"]]) * sqrt(scale[["overlap"]])
```

# Greeks

```{r}
level_shock <- function(shock, S, tau, sigma) {
    
    result <- S * (1 + shock * sigma * sqrt(tau))
    
    return(result)
    
}
```

```{r}
# https://en.wikipedia.org/wiki/Greeks_(finance)
# https://www.wolframalpha.com/input/?i=option+pricing+formula
factor <- "SP500"
types <- c("call", "put")
S <- coredata(na.locf(levels_xts[nrow(levels_xts), factor]))
K <- S
r <- 0 # use "USD3MTD156N"
q <- 0 # see https://stackoverflow.com/a/11286679 
tau <- 1 # = 252 / 252
sigma <- coredata(sd_xts[nrow(sd_xts), factor]) # use "VIXCLS"
shocks <- seq(-3, 3, by = 0.5)
```

```{r}
greeks_dt <- CJ(type = types, shock = shocks)
greeks_dt[ , spot := level_shock(shock, S, tau, sigma), by = c("type", "shock")]
```

## Value

For a given spot price $S$, strike price $K$, risk-free rate $r$, annual dividend yield $q$, time-to-maturity $\tau = T - t$, and volatility $\sigma$:

$$
\begin{aligned}
V_{c}&=Se^{-q\tau}\Phi(d_{1})-e^{-r\tau}K\Phi(d_{2}) \\
V_{p}&=e^{-r\tau}K\Phi(-d_{2})-Se^{-q\tau}\Phi(-d_{1})
\end{aligned}
$$

```{r}
bs_value <- function(type, S, K, r, q, tau, sigma, d1, d2) {
  
    r_df <- exp(-r * tau)
    q_df <- exp(-q * tau)
    
    call_value <- S * q_df * Phi(d1) - r_df * K * Phi(d2)
    put_value <- r_df * K * Phi(-d2) - S * q_df * Phi(-d1)
    result <- ifelse(type == "call", call_value, put_value)
    
    return(result)
    
} 
```

where

$$
\begin{aligned}
d_{1}&={\frac{\ln(S/K)+(r-q+\sigma^{2}/2)\tau}{\sigma{\sqrt{\tau}}}} \\
d_{2}&={\frac{\ln(S/K)+(r-q-\sigma^{2}/2)\tau}{\sigma{\sqrt{\tau}}}}=d_{1}-\sigma{\sqrt{\tau}} \\
\phi(x)&={\frac{e^{-{\frac {x^{2}}{2}}}}{\sqrt{2\pi}}} \\
\Phi(x)&={\frac{1}{\sqrt{2\pi}}}\int_{-\infty}^{x}e^{-{\frac{y^{2}}{2}}}dy=1-{\frac{1}{\sqrt{2\pi}}}\int_{x}^{\infty}e^{-{\frac{y^{2}}{2}}dy}
\end{aligned}
$$

```{r}
bs_d1 <- function(S, K, r, q, tau, sigma) {
    
    result <- (log(S / K) + (r - q + sigma ^ 2 / 2) * tau) / (sigma * sqrt(tau))
    
    return(result)
    
}

bs_d2 <- function(S, K, r, q, tau, sigma) {
    
    result <- (log(S / K) + (r - q - sigma ^ 2 / 2) * tau) / (sigma * sqrt(tau))
    
    return(result)
    
}

phi <- function(x) {
    
    result <- dnorm(x)
    
    return(result)
    
}

Phi <- function(x) {
    
    result <- pnorm(x)
    
    return(result)
    
}
```

```{r}
greeks_dt[ , d1 := bs_d1(spot, K, r, q, tau, sigma), by = c("type", "shock")]
greeks_dt[ , d2 := bs_d2(spot, K, r, q, tau, sigma), by = c("type", "shock")]
greeks_dt[ , value := bs_value(type, spot, K, r, q, tau, sigma, d1, d2), by = c("type", "shock")]
```

## First-order

### Delta

$$
\begin{aligned}
\Delta_{c}&={\frac{\partial V_{c}}{\partial S}}=e^{-q\tau}\Phi(d_{1}) \\
\Delta_{p}&={\frac{\partial V_{p}}{\partial S}}=-e^{-q\tau}\Phi(-d_{1})
\end{aligned}
$$

```{r}
bs_delta <- function(type, S, K, r, q, tau, sigma, d1, d2) {
    
    q_df <- exp(-q * tau)
  
    call_value <- q_df * Phi(d1)
    put_value <- -q_df * Phi(-d1)
    result <- ifelse(type == "call", call_value, put_value)
    
    return(result)
    
} 
```

```{r}
greeks_dt[ , delta := bs_delta(type, spot, K, r, q, tau, sigma, d1, d2), by = c("type", "shock")]
```

### Vega

$$
\begin{aligned}
\nu_{c,p}&={\frac{\partial V_{c,p}}{\partial\sigma}}=Se^{-q\tau}\phi(d_{1}){\sqrt{\tau}}=Ke^{-r\tau}\phi(d_{2}){\sqrt{\tau}}
\end{aligned}
$$

```{r}
bs_vega <- function(type, S, K, r, q, tau, sigma, d1, d2) {
    
    q_df <- exp(-q * tau)
    
    result <- S * q_df * phi(d1) * sqrt(tau)
    
    return(result)
    
}
```

```{r}
greeks_dt[ , vega := bs_vega(type, spot, K, r, q, tau, sigma, d1, d2), by = c("type", "shock")]
```

### Theta

$$
\begin{aligned}
\Theta_{c}&=-{\frac{\partial V_{c}}{\partial \tau}}=-e^{-q\tau}{\frac{S\phi(d_{1})\sigma}{2{\sqrt{\tau}}}}-rKe^{-r\tau}\Phi(d_{2})+qSe^{-q\tau}\Phi(d_{1}) \\
\Theta_{p}&=-{\frac{\partial V_{p}}{\partial \tau}}=-e^{-q\tau}{\frac{S\phi(d_{1})\sigma}{2{\sqrt{\tau}}}}+rKe^{-r\tau}\Phi(-d_{2})-qSe^{-q\tau}\Phi(-d_{1})
\end{aligned}
$$

```{r}
bs_theta <- function(type, S, K, r, q, tau, sigma, d1, d2) {
    
    r_df <- exp(r * tau)
    q_df <- exp(q * tau)
  
    call_value <- -q_df * S * phi(d1) * sigma / (2 * sqrt(tau)) -
      r * K * r_df * Phi(d2) + q * S * q_df * Phi(d1)
    
    put_value <- -q_df * S * phi(d1) * sigma / (2 * sqrt(tau)) +
      r * K * r_df * Phi(-d2) - q * S * q_df * Phi(-d1)
        
    result <- ifelse(type == "call", call_value, put_value)
    
    return(result)
    
}
```

```{r}
greeks_dt[ , theta := bs_theta(type, spot, K, r, q, tau, sigma, d1, d2), by = c("type", "shock")]
```

## Second-order

### Gamma

$$
\begin{aligned}
\Gamma_{c,p}&={\frac{\partial\Delta_{c,p}}{\partial S}}={\frac{\partial^{2}V_{c,p}}{\partial S^{2}}}=e^{-q\tau}{\frac{\phi(d_{1})}{S\sigma{\sqrt{\tau}}}}=Ke^{-r\tau}{\frac{\phi(d_{2})}{S^{2}\sigma{\sqrt{\tau}}}}
\end{aligned}
$$

```{r}
bs_gamma <- function(type, S, K, r, q, tau, sigma, d1, d2) {
  
    q_df <- exp(-q * tau)
    
    result <- q_df * phi(d1) / (S * sigma * sqrt(tau))
    
    return(result)
    
}
```

```{r}
greeks_dt[ , gamma := bs_gamma(type, spot, K, r, q, tau, sigma, d1, d2), by = c("type", "shock")]
```

```{r, echo = FALSE, fig.width = 10, fig.height = 3}
greeks_mlt <- copy(greeks_dt)[ , c("spot", "d1", "d2") := NULL]
greeks_mlt <- melt(greeks_mlt, id.vars = c("type", "shock"))
greeks_plt <- plot_scen(greeks_mlt, z = "type", title = "Greeks", xlab = "Shock") +
  facet_wrap(~ variable, nrow = 1, scales = "free", labeller = labeller(variable = capitalize))
print(greeks_plt)
```

## Nonlinear beta

Notional market value is the market value of a leveraged position:

$$
\begin{aligned}
\text{Equity options }=&\,\#\text{ contracts}\times\text{multiple}\times\text{spot price}\\
\text{Delta-adjusted }=&\,\#\text{ contracts}\times\text{multiple}\times\text{spot price}\times\text{delta}
\end{aligned}
$$

```{r}
# https://en.wikipedia.org/wiki/Notional_amount
beta_option <- function(type, S, K, r, q, tau, sigma, sec) {
    
    d1 <- bs_d1(S, K, r, q, tau, sigma)
    d2 <- bs_d2(S, K, r, q, tau, sigma)
    delta <- bs_delta(type, S, K, r, q, tau, sigma, d1, d2)
    
    notional_mv <- sec[["n"]] * sec[["multiple"]] * sec[["S"]] * delta
    notional_mv0 <- sec[["n"]] * sec[["multiple"]] * sec[["S"]] * sec[["delta"]]
    
    if (type == "call") {
        result <- sec[["beta"]] * (notional_mv - notional_mv0)
    } else if (type == "put") {
        result <- sec[["beta"]] * (notional_mv0 - notional_mv)
    }        
        
    return(result)
    
}
```

```{r}
beta <- 0.35
type <- "call"
n <- 200
multiple <- 100
nav <- 1000000000
```

```{r}
d1 <- bs_d1(S, K, r, q, tau, sigma)
d2 <- bs_d2(S, K, r, q, tau, sigma)
delta <- bs_delta(type, S, K, r, q, tau, sigma, d1, d2)
sec <- list(
    "n" = n,
    "multiple" = multiple,
    "S" = S,
    "delta" = delta,
    "beta" = 1
)
```

```{r}
beta_dt <- CJ(type = type, shock = shocks)
beta_dt[ , spot := level_shock(shock, S, tau, sigma), by = c("type", "shock")]
beta_dt[ , static := beta]
beta_dt[ , dynamic := beta + beta_option(type, spot, K, r, q, tau, sigma, sec) / nav, by = c("type", "shock")]
```

```{r, echo = FALSE, fig.width = 4, fig.height = 3}
beta_mlt <- copy(beta_dt)[ , c("type", "spot") := NULL]
beta_mlt <- melt(beta_mlt, id.vars = "shock")
beta_plt <- plot_scen(beta_mlt, title = "Beta", xlab = "Shock")
print(beta_plt)
```

For completeness, duration equivalent is defined as:

$$
\begin{aligned}
\text{10-year equivalent }=\,&\frac{\text{security duration}}{\text{10-year OTR duration}}
\end{aligned}
$$

# Taylor series

## Price-yield formula

For a function of one variable, $f(x)$, the Taylor series formula is:

$$
\begin{aligned}
f(x+\Delta x)&=f(x)+{\frac{f'(x)}{1!}}\Delta x+{\frac{f''(x)}{2!}}(\Delta x)^{2}+{\frac{f^{(3)}(x)}{3!}}(\Delta x)^{3}+\cdots+{\frac{f^{(n)}(x)}{n!}}(\Delta x)^{n}+\cdots\\
f(x+\Delta x)-f(x)&={\frac{f'(x)}{1!}}\Delta x+{\frac{f''(x)}{2!}}(\Delta x)^{2}+{\frac{f^{(3)}(x)}{3!}}(\Delta x)^{3}+\cdots+{\frac{f^{(n)}(x)}{n!}}(\Delta x)^{n}+\cdots
\end{aligned}
$$

Using the price-yield formula, the estimated percentage change in price for a change in yield is:

$$
\begin{aligned}
P(y+\Delta y)-P(y)&\approx{\frac{P'(y)}{1!}}\Delta y+{\frac{P''(y)}{2!}}(\Delta y)^{2}\\
&\approx -D\Delta y +{\frac{C}{2!}}(\Delta y)^{2}
\end{aligned}
$$

```{r}
pnl_bond <- function(duration, convexity, dy) {
    
    duration_pnl <- -duration * dy
    convexity_pnl <- (convexity / 2) * dy ^ 2
    
    result <- list("duration_pnl" = duration_pnl,
                   "convexity_pnl" = convexity_pnl)
    
    return(result)
    
} 
```

```{r}
# https://onlinelibrary.wiley.com/doi/pdf/10.1002/9781118267967.app1
duration <- 6.5 * (2 / 3)
convexity <- 0.65 * (2 / 3)
dy <- 0.0025 # 25 bps
```

```{r, eval = FALSE}
pnl_bond(duration, convexity, dy)
```

### Duration drift

The derivative of duration with respect to interest rates gives:

$$
\begin{aligned}
\text{Drift}&=-\frac{\partial D}{\partial y}\\
&=\frac{1}{P}\frac{\partial^{2}P}{\partial y^{2}}-\frac{1}{P^{2}}\frac{\partial P}{\partial y}\frac{\partial P}{\partial y}\\
&=C-D^{2}
\end{aligned}
$$

Because of market conventions, use the following formula: $\text{Drift}=\frac{1}{100}\left(C\times 100-D^{2}\right)=C-\frac{D^{2}}{100}$

```{r}
yield_shock <- function(shock, tau, sigma) {
    
    result <- shock * sigma * sqrt(tau)
    
    return(result)
    
}
```

```{r}
duration_drift <- function(duration, convexity, dy) {
    
    drift <- convexity - duration ^ 2 / 100
    change <- -drift * dy * 100
    
    result <- list("drift" = drift,
                   "change" = change)
    
    return(result)
    
}
```

```{r}
# "Risk Management: Approaches for Fixed Income Markets" (page 45)
factor <- "DGS10"
sigma <- coredata(sd_xts[nrow(sd_xts), factor])
```

```{r}
duration_dt <- CJ(shock = shocks)
duration_dt[ , spot := yield_shock(shock, tau, sigma), by = "shock"]
duration_dt[ , static := duration]
duration_dt[ , dynamic := duration + duration_drift(duration, convexity, spot)[["change"]], by = "shock"]
```

```{r, fig.width = 4, fig.height = 3}
duration_mlt <- copy(duration_dt)[ , "spot" := NULL]
duration_mlt <- melt(duration_mlt, id.vars = "shock")
duration_plt <- plot_scen(duration_mlt, title = "Duration", xlab = "Shock")
print(duration_plt)
```

## Black's formula

A similar formula holds for functions of several variables $f(x_{1},\ldots,x_{n})$. This is usually written as:

$$
\begin{aligned}
f(x_{1}+\Delta x_{1},\ldots,x_{n}+\Delta x_{n})&=f(x_{1},\ldots, x_{n})+ \sum _{j=1}^{n}{\frac{\partial f(x_{1},\ldots,x_{n})}{\partial x_{j}}}(\Delta x_{j})\\
&+{\frac {1}{2!}}\sum_{j=1}^{n}\sum_{k=1}^{n}{\frac{\partial^{2}f(x_{1},\ldots,x_{d})}{\partial x_{j}\partial x_{k}}}(\Delta x_{j})(\Delta x_{k})+\cdots
\end{aligned}
$$

Using Black's formula, the estimated change of an option price is:

$$
\begin{aligned}
V(S+\Delta S,\sigma+\Delta\sigma,t+\Delta t)-V(S,\sigma,t)&\approx{\frac{\partial V}{\partial S}}\Delta S+{\frac{1}{2!}}{\frac{\partial^{2}V}{\partial S^{2}}}(\Delta S)^{2}+{\frac{\partial V}{\partial \sigma}}\Delta\sigma+{\frac{\partial V}{\partial t}}\Delta t\\
&\approx \Delta_{c,p}\Delta S+{\frac{1}{2!}}\Gamma_{c,p}(\Delta S)^{2}+\nu_{c,p}\Delta\sigma+\Theta_{c,p}\Delta t
\end{aligned}
$$

```{r}
# unable to verify the result
pnl_option <- function(type, S, K, r, q, tau, sigma, dS, dt, dsigma) {
    
    d1 <- bs_d1(S, K, r, q, tau, sigma)
    d2 <- bs_d2(S, K, r, q, tau, sigma)
    value <- bs_value(type, S, K, r, q, tau, sigma, d1, d2)
    delta <- bs_delta(type, S, K, r, q, tau, sigma, d1, d2)
    vega <- bs_vega(type, S, K, r, q, tau, sigma, d1, d2)
    theta <- bs_theta(type, S, K, r, q, tau, sigma, d1, d2)
    gamma <- bs_gamma(type, S, K, r, q, tau, sigma, d1, d2)
    
    delta_pnl <- delta * dS / value
    gamma_pnl <- gamma * dS ^ 2 / value
    vega_pnl <- vega * dsigma / value
    theta_pnl <- theta * dt / value
    
    result <- list("total" = delta_pnl + gamma_pnl + vega_pnl + theta_pnl,
                   "delta" = delta_pnl,
                   "gamma" = gamma_pnl,
                   "vega" = vega_pnl,
                   "theta" = theta_pnl)
    
    return(result)    
    
}
```

```{r}
factor <- "SP500"
type <- "call"
S <- coredata(tail(na.locf(levels_xts[ , factor]), width)[1])
K <- S * (1 + 0.05)
tau <- 1 # = 252 / 252
sigma <- coredata(tail(sd_xts[ , factor], width)[1])
```

```{r}
options_dt <- data.table(index = index(tail(levels_xts, width)),
                         spot = na.locf(tail(levels_xts[ , factor], width)),
                         sigma = tail(sd_xts[ , factor], width))
setnames(options_dt, c("index", "spot", "sigma"))
options_dt[ , dS := spot - S, by = index]
options_dt[ , dt := (index - index(tail(levels_xts, width))[1]) / 360, by = index]
options_dt[ , dsigma := sigma - ..sigma, by = index]
```

```{r}
attrib_dt <- options_dt[ , as.list(unlist(pnl_option(type, S, K, r, q, tau, ..sigma,
                                                     dS, dt, dsigma))), by = index]
```

```{r, fig.width = 6, fig.height = 3}
attrib_mlt <- melt(attrib_dt, id.vars = "index")
attrib_plt <- plot_ts_decomp(attrib_mlt, decomp = "Total", title = "Attribution 1Y (%)")
print(attrib_plt)
```

## Ito's lemma

For a given diffiusion $X(t, w)$ driven by:

$$
\begin{aligned}
dX_{t}&=\mu_{t}dt+\sigma_{t}dB_{t}
\end{aligned}
$$

Then proceed with the Taylor series for a function of two variables $f(t,x)$:

$$
\begin{aligned}
df&={\frac{\partial f}{\partial t}}dt+{\frac{\partial f}{\partial x}}dx+{\frac{1}{2}}{\frac{\partial^{2}f}{\partial x^{2}}}dx^{2}\\
&={\frac{\partial f}{\partial t}}dt+{\frac{\partial f}{\partial x}}(\mu_{t}dt+\sigma_{t}dB_{t})+{\frac{1}{2}}{\frac{\partial^{2}f}{\partial x^{2}}}\left(\mu_{t}^{2}dt^{2}+2\mu_{t}\sigma _{t}dtdB_{t}+\sigma_{t}^{2}dB_{t}^{2}\right)\\
&=\left({\frac{\partial f}{\partial t}}+\mu_{t}{\frac{\partial f}{\partial x}}+{\frac{\sigma _{t}^{2}}{2}}{\frac{\partial ^{2}f}{\partial x^{2}}}\right)dt+\sigma_{t}{\frac{\partial f}{\partial x}}dB_{t}
\end{aligned}
$$

Note: set the $dt^{2}$ and $dtdB_{t}$ terms to zero and substitute $dt$ for $dB^{2}$.

### Geometric Brownian motion

The most common application of Ito's lemma in finance is to start with the percent change of an asset:

$$
\begin{aligned}
\frac{dS}{S}&=\mu_{t}dt+\sigma_{t}dB_{t}
\end{aligned}
$$

Then apply Ito's lemma with $f(S)=log(S)$:

$$
\begin{aligned}
d\log(S)&=f^{\prime}(S)dS+{\frac{1}{2}}f^{\prime\prime}(S)S^{2}\sigma^{2}dt\\
&={\frac {1}{S}}\left(\sigma SdB+\mu Sdt\right)-{\frac{1}{2}}\sigma^{2}dt\\
&=\sigma dB+\left(\mu-{\tfrac{\sigma^{2}}{2}}\right)dt
\end{aligned}
$$

It follows that:

$$
\begin{aligned}
\log(S_{t})-\log(S_{0})=\sigma dB+\left(\mu-{\tfrac{\sigma^{2}}{2}}\right)dt
\end{aligned}
$$

Exponentiating gives the expression for $S$:

$$
\begin{aligned}
S_{t}=S_{0}\exp\left(\sigma B_{t}+\left(\mu-{\tfrac{\sigma^{2}}{2}}\right)t\right)
\end{aligned}
$$

This provides a recursive procedure for simulating values of $S$ at $t_{0}<t_{1}<\cdots<t_{n}$:

$$
\begin{aligned}
S(t_{i+1})&=S(t_{i})\exp\left(\sigma\sqrt{t_{i+1}-t_{i}}Z_{i+1}+\left[\mu-{\tfrac{\sigma^{2}}{2}}\right]\left(t_{i+1}-t_{i}\right)\right)
\end{aligned}
$$

where $Z_{1},Z_{2},\ldots,Z_{n}$ are independent standard normals.

```{r}
sim_gbm <- function(n_sim, S, mu, sigma, dt) {
    
    result <- S * exp(cumsum(sigma * sqrt(dt) * rnorm(n_sim)) +
                      (mu - 0.5 * sigma ^ 2) * dt)
    
    return(result)
    
}
```

This leads to an algorithm for simulating a multidimensional geometric Brownian motion:

$$
\begin{aligned}
S_{k}(t_{i+1})&=S_{k}(t_{i})\exp\left(\sqrt{t_{i+1}-t_{i}}\sum_{j=1}^{d}{A_{kj}Z_{i+1,j}}+\left[\mu_{k}-{\tfrac{\sigma_{k}^{2}}{2}}\right]\left(t_{i+1}-t_{i}\right)\right)
\end{aligned}
$$

where $A$ is the Cholesky factor of $\Sigma$, i.e. $A$ is any matrix for which $AA^\mathrm{T}=\Sigma$.

```{r}
sim_multi_gbm <- function(n_sim, S, mu, sigma, dt) {
    
    n_cols <- ncol(sigma)
    
    Z <- matrix(rnorm(n_sim * n_cols), nrow = n_sim, ncol = n_cols)
    X <- sweep(sqrt(dt) * (Z %*% chol(sigma)), 2, (mu - 0.5 * diag(sigma)) * dt, "+")
    
    result <- sweep(apply(X, 2, function(x) exp(cumsum(x))), 2, S, "*")
    
    return(result)
    
}
```

```{r}
# https://arxiv.org/pdf/0812.4210.pdf
# https://quant.stackexchange.com/questions/15219/calibration-of-a-gbm-what-should-dt-be
# https://stackoverflow.com/questions/36463227/geometrical-brownian-motion-simulation-in-r
# https://quant.stackexchange.com/questions/25219/simulate-correlated-geometric-brownian-motion-in-the-r-programming-language
# https://quant.stackexchange.com/questions/35194/estimating-the-historical-drift-and-volatility/
S <- rep(1, length(factors))
sigma <- cov(returns_xts, use = "complete.obs") * scale[["periods"]]
mu <- colMeans(na.omit(returns_xts)) * scale[["periods"]]
mu <- mu + diag(sigma) / 2 # drift
dt <- 1 / scale[["periods"]]
```

```{r}
mu_ls <- list()
sigma_ls <- list()
```

```{r}
for (i in 1:1000) {
    
    # assumes stock prices
    levels_sim <- sim_multi_gbm(width + 1, S, mu, sigma, dt)
    returns_sim <- diff(log(levels_sim))

    mu_sim <- colMeans(returns_sim) * scale[["periods"]]
    sigma_sim <- apply(returns_sim, 2, sd) * sqrt(scale[["periods"]])
    
    mu_ls <- append(mu_ls, list(mu_sim))
    sigma_ls <- append(sigma_ls, list(sigma_sim))
    
}
```

```{r}
data.frame("empirical" = colMeans(na.omit(returns_xts)) * scale[["periods"]],
           "theoretical" = colMeans(do.call("rbind", mu_ls)))
```

```{r}
data.frame("empirical" = sqrt(diag(sigma)),
           "theoretical" = colMeans(do.call("rbind", sigma_ls)))
```

# Implied volatility

## Newton's method

Newton's method (main idea is also from a Taylor series) is a method for finding approximations to the roots of a function $f(x)$:

$$
\begin{aligned}
x_{n+1}=x_{n}-{\frac{f(x_{n})}{f'(x_{n})}}
\end{aligned}
$$

To solve $V(\sigma_{n})-V=0$ for $\sigma_{n}$, use Newton's method and repeat until $\left|\sigma_{n+1}-\sigma_{n}\right|<\varepsilon$:

$$
\begin{aligned}
\sigma_{n+1}=\sigma_{n}-{\frac{V(\sigma_{n})-V}{V'(\sigma_{n})}}
\end{aligned}
$$

```{r}
implied_vol_newton <- function(params, type, S, K, r, q, tau) {
    
    target0 <- 0
    sigma0 <- params[["sigma"]]
    
    while (abs(target0 - params[["target"]]) > params[["tol"]]) {
        
        d1 <- bs_d1(S, K, r, q, tau, sigma0)
        d2 <- bs_d2(S, K, r, q, tau, sigma0)
        
        target0 <- bs_value(type, S, K, r, q, tau, sigma0, d1, d2)
        vega0 <- bs_vega(type, S, K, r, q, tau, sigma0, d1, d2)
        
        sigma <- sigma0 - (target0 - params[["target"]]) / vega0
        sigma0 <- sigma
        
    }
    
    return(sigma)
    
}
```

```{r}
# http://www.aspenres.com/documents/help/userguide/help/bopthelp/bopt2Implied_Volatility_Formula.html
# https://books.google.com/books?id=VLi61POD61IC&pg=PA104
S <- coredata(na.locf(levels_xts)[nrow(levels_xts), factor])
K <- S * (1 + 0.05)
sigma <- coredata(sd_xts[nrow(sd_xts), factor]) # overrides matrix
start <- 0.2
```

```{r}
d1 <- bs_d1(S, K, r, q, tau, sigma)
d2 <- bs_d2(S, K, r, q, tau, sigma)
target <- bs_value(type, S, K, r, q, tau, sigma, d1, d2)
params <- list(
    "target" = target,
    "sigma" = start,
    "tol" = 1e-4 # .Machine$double.eps
)
```

```{r}
implied_vol_newton(params, type, S, K, r, q, tau)
```

## Optimization

If the derivative is unknown, try optimization:

```{r}
implied_vol_obj <- function(param, type, S, K, r, q, tau, target) {
    
    d1 <- bs_d1(S, K, r, q, tau, param)
    d2 <- bs_d2(S, K, r, q, tau, param)
    target0 <- bs_value(type, S, K, r, q, tau, param, d1, d2)
    
    result <- abs(target0 - target)
    
    return(result)
    
}

implied_vol_optim <- function(param, type, S, K, r, q, tau, target) {
    
    result <- optim(param, implied_vol_obj, type = type, S = S, K = K, r = r, q = q,
                    tau = tau, target = target, method = "Brent", lower = 0, upper = 1)
    
    return(result$par)
    
}
```

```{r}
implied_vol_optim(start, type, S, K, r, q, tau, target)
```

## Variance swaps

A variance swap can be written as:

$$
\begin{aligned}
K_{var}={\frac{2e^{rT}}{T}}\left(\int\limits_{0}^{F_{0}}{\frac{1}{K^{2}}}P(K)dK+\int\limits_{F_{0}}^{\infty}{\frac{1}{K^{2}}} C(K)dK\right)
\end{aligned}
$$

The CBOE Volatility Index (VIX) is calculated as a variance swap on the 30-day variance of the S&P 500 with an adjustment term:

$$
\begin{aligned}
\sigma^{2}={\frac{2e^{rT}}{T}}\left(\sum_{i=0}^{K_{0}}\frac{\Delta K_{i}}{K_{i}^{2}}P(K_{i})+\sum_{i=K_{0}}^{\infty}\frac{\Delta K_{i}}{K_{i}^{2}}C(K_{i})\right)-\frac{1}{T}\left(\frac{F}{K_{0}}-1\right)^{2}
\end{aligned}
$$

```{r}
# https://cdn.cboe.com/resources/vix/vixwhite.pdf
# https://en.wikipedia.org/wiki/VIX
# https://en.wikipedia.org/wiki/Variance_swap
# https://www.ivolatility.com/doc/VarianceSwaps.pdf
implied_vol_vix <- function(calls_df, puts_df, r, tau) {
  
  # time to expiration  
  t <- as.numeric(difftime(tau[["exp_time"]], tau[["sys_time"]], units = "mins")) / (365 * 24 * 60)
  
  # midpoint of bid and ask
  calls_df[["Mid"]] <- rowMeans(calls_df[ , c("Bid", "Ask")])
  puts_df[["Mid"]] <- rowMeans(puts_df[ , c("Bid", "Ask")])
  
  options_df <- merge(calls_df[ , c("Strike", "Mid")],
                      puts_df[ , c("Strike", "Mid")], by = "Strike")
  colnames(options_df) <- c("Strike", "Call", "Put")
  
  options_df[["Diff"]] <- abs(options_df[["Call"]] - options_df[["Put"]])
  
  # minimum absolute difference is forward index level
  forward_df <- options_df[options_df[["Diff"]] == min(options_df[["Diff"]]), ]
  k <- forward_df[["Strike"]]
  c <- forward_df[["Call"]]
  p <- forward_df[["Put"]]
  f <- k + exp(r * t) * (c - p)
  
  # strike price equal or below forward index level
  k0 <- tail(options_df[options_df[["Strike"]] <= f, "Strike"], 1)
  
  # out-of-the-money options
  puts_otm_df <- puts_df[puts_df[["Strike"]] < k0, ]
  calls_otm_df <- calls_df[calls_df[["Strike"]] > k0, ]
  
  # stop after two consecutive strike prices with zero bid prices
  # https://stackoverflow.com/a/50311890
  puts_otm_rle <- rle(puts_otm_df[["Bid"]])
  idx <- cumsum(puts_otm_rle$lengths) # end
  idx <- idx[max(which(puts_otm_rle$lengths > 1 & puts_otm_rle$values == 0))]
  puts_otm_df <- puts_otm_df[(idx + 1):nrow(puts_otm_df), ]
  
  calls_otm_rle <- rle(calls_otm_df[["Bid"]])
  idx <- cumsum(calls_otm_rle$lengths) # end
  idx <- idx - calls_otm_rle$lengths + 1 # start
  idx <- idx[max(which(calls_otm_rle$lengths > 1 & calls_otm_rle$values == 0))]
  calls_otm_df <- calls_otm_df[1:(idx - 1), ]
  
  # average put and call prices for k0
  # note: exclude options with zero bid price
  result_df <- rbind(puts_otm_df[puts_otm_df[["Bid"]] != 0, c("Strike", "Mid")],
                     colMeans(rbind(puts_df[puts_df[["Strike"]] == k0, c("Strike", "Mid")],
                                    calls_df[calls_df[["Strike"]] == k0, c("Strike", "Mid")])),
                     calls_otm_df[calls_otm_df[["Bid"]] != 0, c("Strike", "Mid")])
  
  # differences between strike prices
  n_rows <- nrow(result_df)
  result_df[1, "Diff"] <- result_df[2, "Strike"] - result_df[1, "Strike"]
  result_df[n_rows, "Diff"] <- result_df[n_rows, "Strike"] - result_df[n_rows - 1, "Strike"]
  result_df[2:(n_rows - 1), "Diff"] <- (result_df[3:n_rows, "Strike"] -
                                          result_df[1:(n_rows - 2), "Strike"]) / 2
  
  # variance
  v <- sum(result_df[ , "Diff"] / result_df[ , "Strike"] ^ 2 * exp(r * t) *
             result_df[ , "Mid"]) * (2 / t) - (1 / t) * (f / k0 - 1) ^ 2
  
  result <- list("t" = t,
                 "v" = v)
  
  return(result)
  
}
```

```{r}
sys_time <- format(Sys.Date(), paste0("%Y-%m-%d", "09:46:00"))
```

```{r}
v1 <- implied_vol_vix(read.csv("../securities/near_calls.csv"), read.csv("../securities/near_puts.csv"), 0.000305,
                      list("sys_time" = sys_time,
                           "exp_time" = format(Sys.Date() + 25, paste0("%Y-%m-%d", "08:30:00"))))

v2 <- implied_vol_vix(read.csv("../securities/next_calls.csv"), read.csv("../securities/next_puts.csv"), 0.000286,
                      list("sys_time" = sys_time,
                           "exp_time" = format(Sys.Date() + 32, paste0("%Y-%m-%d", "15:00:00"))))
```

```{r}
nt1 <- v1$t * (365 * 24 * 60)
nt2 <- v2$t * (365 * 24 * 60)
n30 <- 30 * (24 * 60)
n365 <- 365 * (24 * 60)
```

```{r}
vix <- sqrt((v1$t * v1$v * ((nt2 - n30) / (nt2 - nt1)) +
               v2$t * v2$v * ((n30 - nt1) / (nt2 - nt1))) * (n365 / n30))
print(vix)
```

## Implied yield

```{r}
yield_option <- function(type, S, K, r, q, tau, sigma) {
    
    d1 <- bs_d1(S, K, r, q, tau, sigma)
    d2 <- bs_d2(S, K, r, q, tau, sigma)
    value <- bs_value(type, S, K, r, q, tau, sigma, d1, d2)
    
    if (type == "call") {
        result <- (value / S) / tau
    } else if (type == "put") {
        result <- (value / K) / tau
    }
    
    return(result)
    
}
```

```{r}
sigmas <- seq(0.1, 0.3, by = 0.04)
taus <- seq(20, 126, by = 20) / 252
```

```{r}
yield_dt <- CJ(sigma = sigmas, tau = taus)
yield_dt[ , yield := yield_option(type, S, K, r, q, tau, sigma), by = c("sigma", "tau")]
yield_dt[ , sigma := sigma * 100]
yield_dt[ , tau := tau * scale[["periods"]]]
```

```{r, fig.width = 4, fig.height = 3}
yield_plt <- plot_heatmap(yield_dt, x = "tau", y = "sigma", z = "yield",
                          title = "Yield (%)", xlab = "Tau", ylab = "Sigma")
print(yield_plt)
```
