---
title: "Software"
author: "[Jason Foster](mailto:jason.j.foster@gmail.com)"
date: last-modified
categories:
  - development
  - r
  - python
execute:
  eval: false
draft: true
---

# System setup

Set environment variable: `PATH=c:\rtools<123>\usr\bin`

# Git and GitHub

```         
git config --global user.name "<NAME>"
git config --global user.email "<EMAIL>"
```

If needed, generate an SSH key: `RStudio > Tools > Git/SVN > Create RSA Key...`

Then give GitHub your SSH public key: `GitHub > SSH and GPG Keys > New SSH key`

Create a new repository on the command line

```         
echo "# <REPO>" >> README.md
git init
git add README.md
git commit -m "first commit"
git branch -M main
git remote add origin git@github.com:<OWNER>/<REPO>.git
git push -u origin main
```

...or push an existing repository from the command line

```         
git remote add origin git@github.com:<OWNER>/<REPO>.git
git branch -M main
git push -u origin main
```

```         
# git remote set-url origin git@github.com:<OWNER>/<REPO>.git
```

# Development workflow

```{r}
RcppArmadillo::RcppArmadillo.package.skeleton(name = "anRpackage",
                                              path = ".",
                                              example_code = FALSE)
```

# Compiled code

```{r}
Rcpp::compileAttributes(verbose = TRUE)
tools::package_native_routine_registration_skeleton(".", character_only = FALSE)
```

# Object documentation

```{r}
roxygen2::roxygenize()
```

# Automated checking

```{r}
# rename R-CMD-check.yaml to check-standard.yaml
usethis::use_github_action("check-standard")
usethis::use_github_action("test-coverage")
```

-   <https://docs.codecov.com/docs/adding-the-codecov-token>

1.  Codecov.io: `Settings > General > Tokens > CODECOV_TOKEN`

2.  GitHub: `Settings > Secrets and variables > Actions > New repository secret > CODECOV_TOKEN`

# Python

-   <https://www.python.org/ftp/python/>

```         
python-3.10.11-amd64.exe
```

Check box for `Add python.exe to PATH`

## Quarto

```         
pip install pyyaml jupyter
```

## Issues

-   <https://github.com/osqp/osqp/issues/385>

Use python 3.10 for `qdldl`

-   <https://github.com/pydata/pandas-datareader/issues/965>

```         
pip install git+https://github.com/pydata/pandas-datareader.git
```
